---
title: "Intership project"
output: html_document
date: "2025-09-27"
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,        
  eval = TRUE,         
  cache = FALSE,
  warning = FALSE,     
  message = FALSE,    
  results = 'hide'     
)
options(width = 80, digits = 3)


library(dplyr)
library(tidyverse)
library(here)
library(galah)
library(sf)
library(spData)
library(ggplot2)
library(Cairo)
library(kableExtra)
library(bookdown)
library(tibble)
library(purrr)
library(knitr)
library(readr)
library(stringr)
library(purrr)
library(htmltools)
```





#### Data description

The ‘Cafes and Restaurants with Seating Capacity’ dataset is sourced from the City of Melbourne’s Census of Land Use and Employment (CLUE). It provides annual records of business establishments located within the Melbourne local government area from 2002 to 2023. The data can be retrieved via the [City of Melbourne Open Data Portal](https://data.melbourne.vic.gov.au/explore/dataset/cafes-and-restaurants-with-seating-capacity/information).


<div style="font-size: 80%;">
```{r var-types,message = FALSE, warning = FALSE, results = 'show'}
# Read the cleaned CSV files into R environment
df <- read_csv("data/cafes-and-restaurants-with-seating-capacity.csv")

df<- df%>%
  filter(
    `Industry (ANZSIC4) description` %in% c(
      "Cafes and Restaurants",
      "Takeaway Food Services"
    )
  )

# Get column names and types
col_types <- map_chr(df, ~ class(.x)[1])
# Split into two groups
first_half <- col_types[1:7]
second_half <- col_types[8:15]
# Build a 2-row tibble with padding between groups
df_split_summary <- tibble::tibble(
  !!!setNames(as.list(first_half), names(first_half)),
  !!!setNames(as.list(second_half), names(second_half))
)
# Use HTML-based styling but ensure it's compatible with PDF conversion
kbl(df_split_summary, format = "html", caption = "Variable Types in CLUE Dataset", 
    table.attr = 'id="tab:var-types"') %>% # Add explicit ID
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"), 
                full_width = FALSE, 
                position = "center")
```
</div>


The Table \@ref(tab:var-types) shows the column names and data types in the dataset. It includes numbers, text, and location data . This mix of data types makes it useful for different kinds of analysis. The variable names in the dataset are generally clear and descriptive.






<div style="font-size: 80%;">
```{r  summary, message = FALSE, warning = FALSE, results = 'show'}


# Dimensions
n_rows <- nrow(df)
n_cols <- ncol(df)

# Duplicates
n_duplicates <- sum(duplicated(df))

# Missing values
missing_summary <- colSums(is.na(df))
total_missing <- sum(missing_summary)
missing_cols <- names(missing_summary[missing_summary > 0])
missing_cols_str <- paste(missing_cols, collapse = ", ")

# Negative or zero seating values
invalid_seating <- sum(df$`Number of seats` < 0, na.rm = TRUE)

# Create horizontal summary table
df_summary <- tibble(
  `Number of Rows` = n_rows,
  `Number of Columns` = n_cols,
  `Number of Duplicate Rows` = n_duplicates,
  `Total Missing Values` = total_missing,
  `Columns with NA Values` = length(missing_cols),
  `Names of Columns with NA Values` = missing_cols_str,
  `Non-positive Seating Values` = invalid_seating
)

# Display the summary table with border and proper alignment
kable(df_summary, format = "html", caption = "Summary of Dataset Structure and Quality Checks") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"), full_width = F, position = "center")



```
</div>





As shown in Table \@ref(tab:summary), the dataset has 63,121 rows and 15 columns, with no duplicate entries and negative seating number. There are 1,581 missing values, all in the location-related columns (Longitude, Latitude, location). 


<div style="font-size: 80%;">
```{r year, message = FALSE, warning = FALSE, results = 'show'}

library(tibble)
library(knitr)

# Get sorted unique census years
census_years <- sort(unique(df$`Census year`))

# Create one-row, one-cell table
census_table <- tibble(`Census Years` = paste(census_years, collapse = ", "))


kable(census_table, format = "html", caption = "Unique Census Years in the Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"), full_width = F)

```
</div>

Table \@ref(tab:year) shows the dataset covers 22 unique census years, ranging from 2002 to 2023. 


<div style="font-size: 80%;">
```{r  type, message = FALSE, warning = FALSE, results = 'show'}
library(dplyr)
library(knitr)
library(kableExtra)

library(dplyr)
library(knitr)
library(kableExtra)

# Get unique values
seating_types <- sort(unique(na.omit(df$`Seating type`)))
areas <- sort(unique(na.omit(df$`CLUE small area`)))

# Pad the shorter list so both have equal length
max_length <- max(length(seating_types), length(areas))
seating_types <- c(seating_types, rep("", max_length - length(seating_types)))
areas <- c(areas, rep("", max_length - length(areas)))

# Combine into a transposed table
transposed_table <- rbind(seating_types, areas)
rownames(transposed_table) <- c("Seating Type", "CLUE Small Area")

# Display as wide table

kable(transposed_table, format = "html", caption = "Unique Values of Seating Type and CLUE Small Area") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"), full_width = F, position = "center")

```
</div>


As we can see from Table \@ref(tab:type), there are two types of seating group, with options like Indoor and Outdoor. The CLUE small area column shows many different areas in Melbourne, this allows us to compare business patterns by area.



```{r seating-boxplot-compact, fig.align = "center",fig.width=8, fig.height=2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = " Distribution of Seating Capacity"}


library(ggplot2)

ggplot(df, aes(y = `Number of seats`)) +
  geom_boxplot(outlier.size = 0.8, fill = "skyblue") +
  coord_cartesian(ylim = c(0, 500)) +  # Zoom in to ignore extreme outliers
  labs(y = "Number of Seats") +
  theme_minimal(base_size = 10)



```

Number of Seats is the only continuous numerical variable available in our dataset for analysis. The Figure \@ref(fig:seating-boxplot-compact) shows that most businesses have fewer than 100 seats, but there are some with very high capacities. These large values are likely from big venues like stadiums or racecourses. The plot helps focus on typical businesses by zooming in and excluding extreme values.


```{r}


# Count occurrences of each industry type
industry_counts <- table(df$`Industry (ANZSIC4) description`)

# Sorted by frequency
sorted_industry_counts <- sort(industry_counts, decreasing = TRUE)


```




<div style="font-size: 80%;">
```{r  ind, message = FALSE, warning = FALSE, results = 'show'}
library(knitr)

# Convert and prepare data frame
industry_df <- as.data.frame(sorted_industry_counts)
colnames(industry_df) <- c("Industry_Type", "Count")



kable(head(industry_df, 8), format = "html", caption = "Table: Top 8 Industry Types by Number of Businesses") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"), full_width = F, position = "center")

```
</div>

The Table \@ref(tab:ind) shows the top 8 industry types in Melbourne’s hospitality sector based on the number of businesses recorded in the dataset. Cafes and Restaurants dominate the landscape with over 47,000 entries, followed by Takeaway Food Services and Pubs. Smaller but still notable categories include Accommodation, Bakeries, and Catering Services, highlighting the variety within the food and hospitality industry across Melbourne.






##### Foot traffic data

```{r}
# Read the cleaned CSV files into R environment
ped <- read_csv("data/foot.csv")

ped <- ped %>%
  mutate(
    date = ymd(Sensing_Date),
    year = year(date),
    # Location = "lat, lon" -> split to numeric lat/lon
    lat = as.numeric(str_trim(str_extract(Location, "^-?\\d+\\.\\d+"))),
    lon = as.numeric(str_trim(str_extract(Location, "-?\\d+\\.\\d+$")))
  )

# --- 3) Aggregate hourly -> daily totals per sensor ----
# Total_of_Directions is the hourly count; sum to daily
daily <- ped %>%
  group_by(Location_ID, Sensor_Name, lat, lon, date, year) %>%
  summarise(
    daily_total = sum(Total_of_Directions, na.rm = TRUE),
    .groups = "drop"
  )

# (Optional sanity check: number of hourly rows per day)
# ped %>% count(Location_ID, date) %>% summary()

# --- 4) Yearly stats per sensor ----
# We compute:
#   - n_days: number of days seen in that year for that sensor
#   - avg_daily: mean of daily totals across observed days
#   - total_year: sum of daily totals (across observed days)
yearly <- daily %>%
  group_by(Location_ID, Sensor_Name, lat, lon, year) %>%
  summarise(
    n_days     = n_distinct(date),
    avg_daily  = mean(daily_total, na.rm = TRUE),
    total_year = sum(daily_total, na.rm = TRUE),
    .groups = "drop"
  )

# Optional: filter out sensors with very few observed days in a year (e.g., < 200)
# yearly <- yearly %>% filter(n_days >= 200)

# --- 5) Also compute a multi-year average per sensor (for a single “overview” map) ----
# This helps if you want one value per sensor across all years available.
overview <- yearly %>%
  group_by(Location_ID, Sensor_Name, lat, lon) %>%
  summarise(
    years_observed    = n(),
    avg_daily_overall = mean(avg_daily, na.rm = TRUE),
    total_all_years   = sum(total_year, na.rm = TRUE),
    .groups = "drop"
  )

# --- 6) Write outputs ----
write_csv(yearly,  "yearly_sensor_traffic.csv")
write_csv(overview, "overview_sensor_traffic.csv")

message("Saved: yearly_sensor_traffic.csv (per sensor x year)")
message("Saved: overview_sensor_traffic.csv (per sensor overall average)")

# --- 7) Quick interactive map (overview average) ----
# Binning for color scale
pal <- colorBin("YlOrRd", domain = overview$avg_daily_overall, bins = 7, na.color = "#CCCCCC")

leaf <- leaflet(overview) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    radius = ~rescale(avg_daily_overall, to = c(3, 16)),
    color = ~pal(avg_daily_overall),
    stroke = FALSE, fillOpacity = 0.8,
    popup = ~paste0(
      "<b>", htmlEscape(Sensor_Name), "</b><br/>",
      "Avg daily (overall): ", comma(round(avg_daily_overall, 0)), "<br/>",
      "Years observed: ", years_observed
    )
  ) |>
  addLegend(
    "bottomright", pal = pal, values = ~avg_daily_overall,
    title = "Avg daily pedestrians<br/>(overall)", opacity = 0.9
  )


```


```{r var-types,message = FALSE, warning = FALSE, results = 'show'}

library(readr)
library(kableExtra)
library(pillar)

# load
df <- read_csv("overview_sensor_traffic.csv", show_col_types = FALSE)

# one row with types; headers are the column names
types <- vapply(df, function(x) pillar::type_sum(x), character(1))

tbl <- as.data.frame(t(types), stringsAsFactors = FALSE)
colnames(tbl) <- names(df)

kbl(
  tbl,
  format = "html",
  caption = "Variable Types in Sensor Traffic Dataset",
  table.attr = 'id="tab:var-types-sensor"'
) |>
  kable_styling(
    bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = TRUE,
    position = "center"
  )


```




```{r}
library(readr)
library(dplyr)
library(kableExtra)

# --- Load data ---
df <- read_csv("overview_sensor_traffic.csv", show_col_types = FALSE)

# --- Dataset structure summary ---
n_rows <- nrow(df)
n_cols <- ncol(df)
n_duplicates <- sum(duplicated(df))

# --- Missing values ---
missing_summary <- colSums(is.na(df))
total_missing <- sum(missing_summary)
missing_cols <- names(missing_summary[missing_summary > 0])
missing_cols_str <- ifelse(length(missing_cols) > 0, paste(missing_cols, collapse = ", "), "None")

# --- Check for invalid numeric values (example: negative or zero totals) ---
invalid_values <- if ("avg_daily_overall" %in% names(df)) {
  sum(df$avg_daily_overall <= 0, na.rm = TRUE)
} else {
  0
}

# --- Create summary table ---
df_summary <- tibble(
  `Number of Rows` = n_rows,
  `Number of Columns` = n_cols,
  `Number of Duplicate Rows` = n_duplicates,
  `Total Missing Values` = total_missing,
  `Columns with NA Values` = length(missing_cols),
  `Names of Columns with NA Values` = missing_cols_str,
  `Non-positive Daily Values` = invalid_values
)

# --- Render the summary table ---
kbl(
  df_summary,
  format = "html",
  caption = "Summary of Sensor Traffic Dataset Structure and Quality Checks"
) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered", "condensed"),
    full_width = FALSE,
    position = "center"
  )

```








```{r seating-boxplot-compact, fig.align = "center",fig.width=8, fig.height=2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = " Distribution of Seating Capacity"}




library(ggplot2)

ggplot(df, aes(y = avg_daily_overall)) +
  geom_boxplot(outlier.size = 0.8, fill = "lightcoral") +
  coord_cartesian(ylim = c(0, quantile(df$avg_daily_overall, 0.99, na.rm = TRUE))) +  # Zoom to exclude extreme outliers
  labs(y = "Average Daily Foot Traffic") +
  theme_minimal(base_size = 10)




```












```{r}
# Simple leaflet map from your 'overview' data
# requires: overview with columns: lat, lon, Sensor_Name, years_observed, avg_daily_overall

suppressPackageStartupMessages({
  library(dplyr)
  library(leaflet)
  library(scales)
  library(htmltools)  # for htmlEscape
})

# keep only rows with valid coords
ov <- overview %>% filter(!is.na(lat), !is.na(lon))

# color scale by average daily pedestrians (multi-year)
pal <- colorBin("YlOrRd", domain = ov$avg_daily_overall, bins = 7, na.color = "#CCCCCC")

# build popup text once (avoids referencing missing columns)
ov <- ov %>%
  mutate(
    popup_text = paste0(
      "<b>", htmlEscape(Sensor_Name), "</b><br/>",
      "Avg daily pedestrians: ", comma(round(avg_daily_overall)), "<br/>",
      "Years observed: ", years_observed
    )
  )

m <- leaflet(ov) |>
  addProviderTiles("CartoDB.Positron") |>   # <--- cleaner background
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    radius = ~pmax(3, rescale(avg_daily_overall, to = c(3, 16))),
    color  = ~pal(avg_daily_overall),
    stroke = FALSE, fillOpacity = 0.85,
    popup = ~popup_text
  ) |>
  addLegend("bottomright", pal = pal, values = ~avg_daily_overall,
            title = "Avg daily pedestrians", opacity = 0.9)

m


```







#### Data transforming




```{r}






# Read raw data
raw_df <- read_csv("data/cafes-and-restaurants-with-seating-capacity.csv")


# Select industry




df_cleaned <- raw_df %>%
  filter(
    `Industry (ANZSIC4) description` %in% c(
      "Cafes and Restaurants",
      "Takeaway Food Services"
    )
  ) %>%
  filter(
    !is.na(`Trading name`),
    !is.na(`Property ID`),
    !is.na(`Census year`),
    str_trim(`Trading name`) != "",
    `Census year` >= 2002,
    `Census year` <= 2023
  ) %>%
  mutate(
    industry = case_when(
      `Industry (ANZSIC4) description` == "Cafes and Restaurants" ~ "Cafes",
      `Industry (ANZSIC4) description` == "Takeaway Food Services" ~ "Takeaway",
      TRUE ~ "Other"
    ),
    trading_name_original = str_trim(str_to_title(`Trading name`)),
    `Number of seats` = ifelse(is.na(`Number of seats`) | `Number of seats` <= 0, 1, `Number of seats`)
  )



# Show industry breakdown
industry_counts <- df_cleaned %>% count(industry)

print(industry_counts)



#  First Word Normalization function


get_normalized_first_word <- function(name) {
  if (is.na(name) || name == "") return(NA)
  
  # Convert to lowercase
  name <- str_to_lower(name)
  
  # Remove ALL punctuation
  name <- str_replace_all(name, "[^a-z0-9\\s]", " ")
  
  # Collapse multiple spaces
  name <- str_squish(name)
  
  # Split into words
  words <- str_split(name, "\\s+")[[1]]
  words <- words[words != ""]
  
  # Remove very short words
  words <- words[nchar(words) >= 3]
  
  if (length(words) == 0) return(NA)
  
  # Get first word
  first_word <- words[1]
  
  # Normalize singular/plural (remove trailing 's' from 5+ char words)
  if (nchar(first_word) >= 5 && str_ends(first_word, "s")) {
    singular <- str_sub(first_word, 1, -2)
    if (nchar(singular) >= 4) {
      first_word <- singular
    }
  }
  
  # Handle common variations
  first_word <- case_when(
    first_word %in% c("madam", "madame") ~ "madam",
    first_word %in% c("cafe", "caffee", "caffe") ~ "cafe",
    TRUE ~ first_word
  )
  
  return(first_word)
}


# Apply Normalization 




df_normalized <- df_cleaned %>%
  mutate(
    normalized_first_word = sapply(`Trading name`, get_normalized_first_word)
  ) %>%
  filter(!is.na(normalized_first_word), nchar(normalized_first_word) >= 2)




#  Consolidate seating types



df_consolidated <- df_normalized %>%
  group_by(`Census year`, `Property ID`, normalized_first_word) %>%
  summarise(
    trading_name = names(sort(table(trading_name_original), decreasing = TRUE))[1],
    building_address = first(`Building address`),
    business_address = first(`Business address`),
    clue_area = first(`CLUE small area`),
    industry = first(industry),
    lon = first(`Longitude`),
    lat = first(`Latitude`),
    seating_type = case_when(
      n_distinct(`Seating type`, na.rm = TRUE) == 1 ~ first(`Seating type`),
      TRUE ~ "Mixed"
    ),
    num_seats = sum(`Number of seats`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(
    census_year = `Census year`,
    property_id = `Property ID`
  )

cat("After seating consolidation:", nrow(df_consolidated), "records\n")



business_groups <- df_consolidated %>%
  group_by(property_id, normalized_first_word) %>%
  mutate(business_id = cur_group_id()) %>%
  ungroup()



# Add tempral metrics




df_with_lifespans <- business_groups %>%
  group_by(business_id) %>%
  arrange(census_year) %>%
  mutate(
    lifespan_years = max(census_year) - min(census_year) + 1,
    lifespan_group = case_when(
      lifespan_years <= 2 ~ "1-2 yrs",
      lifespan_years <= 5 ~ "3-5 yrs", 
      lifespan_years <= 8 ~ "6-8 yrs",
      lifespan_years <= 12 ~ "9-12 yrs",
      TRUE ~ ">12 yrs"
    )
  ) %>%
  ungroup()

# Add property metrics
property_counts <- df_with_lifespans %>%
  group_by(property_id, census_year) %>%
  summarise(businesses_at_property = n_distinct(business_id), .groups = "drop")

df_final <- df_with_lifespans %>%
  left_join(property_counts, by = c("property_id", "census_year")) %>%
  select(
    property_id, business_id, census_year, industry, clue_area, 
    trading_name, business_address, building_address, seating_type, 
    num_seats, lifespan_years, lifespan_group, lon, lat, businesses_at_property
  ) %>%
  arrange(census_year, property_id, business_id)



#  Fix remaining name variations



# Find potential name variations at same property
property_businesses <- df_final %>%
  group_by(property_id, business_id) %>%
  summarise(
    trading_name = first(trading_name),
    years = paste(min(census_year), max(census_year), sep = "-"),
    .groups = "drop"
  ) %>%
  group_by(property_id) %>%
  mutate(
    businesses_at_property = n(),
    name_simple = str_to_lower(str_remove_all(trading_name, "[^a-z0-9]"))
  ) %>%
  ungroup()

# Create merge map for similar names
merge_map <- list()
merge_count <- 0

properties_to_check <- unique(property_businesses$property_id[property_businesses$businesses_at_property >= 2])

for (prop_id in properties_to_check) {
  businesses <- property_businesses %>% filter(property_id == prop_id)
  
  if (nrow(businesses) < 2) next
  
  for (i in 1:(nrow(businesses) - 1)) {
    for (j in (i + 1):nrow(businesses)) {
      name1 <- businesses$name_simple[i]
      name2 <- businesses$name_simple[j]
      
      # Remove "the" prefix for comparison
      clean1 <- str_remove(name1, "^the")
      clean2 <- str_remove(name2, "^the")
      
      # Check if they should merge
      should_merge <- FALSE
      
      # Rule 1: Exact match after removing "the"
      if (clean1 == clean2) {
        should_merge <- TRUE
      }
      
      # Rule 2: One is subset with small difference
      else if (nchar(clean1) >= 5 && nchar(clean2) >= 5) {
        if (str_detect(clean1, fixed(clean2)) || str_detect(clean2, fixed(clean1))) {
          len_diff <- abs(nchar(clean1) - nchar(clean2))
          if (len_diff <= 10) {
            should_merge <- TRUE
          }
        }
      }
      
      if (should_merge) {
        id1 <- businesses$business_id[i]
        id2 <- businesses$business_id[j]
        
        # Merge to smaller ID
        if (id1 < id2) {
          merge_map[[as.character(id2)]] <- id1
        } else {
          merge_map[[as.character(id1)]] <- id2
        }
        merge_count <- merge_count + 1
      }
    }
  }
}



# Apply merges
get_final_id <- function(id) {
  while (!is.null(merge_map[[as.character(id)]])) {
    id <- merge_map[[as.character(id)]]
  }
  return(id)
}

df_ultra_final <- df_final %>%
  mutate(business_id = sapply(business_id, get_final_id))

# Recalculate metrics after merges
df_ultra_final <- df_ultra_final %>%
  group_by(business_id) %>%
  mutate(
    lifespan_years = max(census_year) - min(census_year) + 1,
    lifespan_group = case_when(
      lifespan_years <= 2 ~ "1-2 yrs",
      lifespan_years <= 5 ~ "3-5 yrs", 
      lifespan_years <= 8 ~ "6-8 yrs",
      lifespan_years <= 12 ~ "9-12 yrs",
      TRUE ~ ">12 yrs"
    )
  ) %>%
  ungroup()

# Recalculate property counts
property_counts_final <- df_ultra_final %>%
  group_by(property_id, census_year) %>%
  summarise(businesses_at_property = n_distinct(business_id), .groups = "drop")

df_ultra_final <- df_ultra_final %>%
  dplyr::select(-businesses_at_property) %>%
  left_join(property_counts_final, by = c("property_id", "census_year")) %>%
  dplyr::select(
    property_id, business_id, census_year, industry, clue_area, 
    trading_name, business_address, building_address, seating_type, 
    num_seats, lifespan_years, lifespan_group, lon, lat, businesses_at_property
  ) %>%
  filter(`num_seats` < 201)%>%
  arrange(census_year, property_id, business_id)




# SAVE 




write_csv(df_ultra_final, "data/df_FINAL_2industries.csv")


# Create business summary
business_summary <- df_ultra_final %>%
  group_by(property_id, business_id) %>%
  summarise(
    trading_name = first(trading_name),
    first_year = min(census_year),
    last_year = max(census_year),
    years_active = n_distinct(census_year),
    industry = first(industry),
    .groups = "drop"
  ) %>%
  mutate(
    year_span = ifelse(first_year == last_year, as.character(first_year), 
                      paste0(first_year, "-", last_year))
  ) %>%
  group_by(property_id) %>%
  mutate(businesses_at_property = n()) %>%
  ungroup() %>%
  dplyr::select(property_id, business_id, trading_name, year_span, years_active, 
         industry, businesses_at_property) %>%
  arrange(property_id, year_span)

write_csv(business_summary, "data/business_summary_ULTRA_FINAL_2industries.csv")


# SUMMARY




# Industry breakdown
industry_final <- df_ultra_final %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  count(industry)


print(industry_final)

# Lifespan distribution
lifespan_dist <- df_ultra_final %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  count(lifespan_group) %>%
  mutate(percentage = round(n / sum(n) * 100, 1))


print(lifespan_dist)


```







```{r}



# Load the Ultra final data
df_ultra <- read_csv("data/df_FINAL_2industries.csv")


# Calculate business operating status 


business_status <- df_ultra %>%
  group_by(business_id) %>%
  summarise(
    business_start_year = min(census_year),
    business_end_year = max(census_year),
    still_operating = (business_end_year == 2023),  
    lifespan_group = first(lifespan_group),
    .groups = "drop"
  )


# ADD bias filters

enhanced_ultra <- df_ultra %>%
  left_join(business_status %>% dplyr::select(business_id, business_start_year, business_end_year, still_operating), 
            by = "business_id") %>%
  mutate(

    # TRUE = exclude from bias-corrected analysis (but keep for temporal analysis)
    # FALSE = include in all analysis
    bias_filter = case_when(
      # Keep ALL >12 year businesses (whether operating or closed)
      lifespan_group == ">12 yrs" ~ FALSE,
      
      # For shorter lifespan groups: only keep closed businesses
      lifespan_group != ">12 yrs" & !still_operating ~ FALSE,  # Closed = keep (true lifespan)
      lifespan_group != ">12 yrs" & still_operating ~ TRUE,    # Still operating = filter out (biased lifespan)
      
      TRUE ~ FALSE
    ),
    
    # Add descriptive status for transparency
    business_status = case_when(
      lifespan_group == ">12 yrs" & still_operating ~ "Long-term Success (Operating)",
      lifespan_group == ">12 yrs" & !still_operating ~ "Long-term Success (Closed)",
      !still_operating ~ "Closed Business",
      still_operating ~ "Still Operating (Biased Lifespan)",
      TRUE ~ "Unknown"
    ),
    
    # Add analysis recommendation
    analysis_usage = case_when(
      bias_filter == FALSE ~ "Use in All Analysis",
      bias_filter == TRUE ~ "Temporal Only (Exclude from Lifespan Analysis)",
      TRUE ~ "Unknown"
    )
  ) %>%
  dplyr::select(-business_start_year, -business_end_year, -still_operating) %>%
  arrange(census_year, property_id, business_id)

# Check quality



# Overall filtering impact
total_records <- nrow(enhanced_ultra)
filtered_records <- sum(enhanced_ultra$bias_filter)
kept_records <- total_records - filtered_records



# Business-level analysis
business_analysis <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  count(lifespan_group, business_status) %>%
  pivot_wider(names_from = business_status, values_from = n, values_fill = 0)


print(business_analysis)

# Lifespan group impact
lifespan_impact <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  group_by(lifespan_group) %>%
  summarise(
    total_businesses = n(),
    filtered_out = sum(bias_filter),
    kept_for_analysis = total_businesses - filtered_out,
    filter_rate = round(filtered_out / total_businesses * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(lifespan_group)


print(lifespan_impact)

# Area impact analysis
area_impact <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  group_by(clue_area) %>%
  summarise(
    total_businesses = n(),
    filtered_out = sum(bias_filter),
    kept_for_analysis = total_businesses - filtered_out,
    filter_rate = round(filtered_out / total_businesses * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(total_businesses))

cat("\n=== Impact by Area (Top 10) ===\n")
print(head(area_impact, 10))

# Temporal impact by year
temporal_impact <- enhanced_ultra %>%
  group_by(census_year) %>%
  summarise(
    total_records = n(),
    filtered_records = sum(bias_filter),
    kept_records = total_records - filtered_records,
    filter_rate = round(filtered_records / total_records * 100, 1),
    .groups = "drop"
  )


print(tail(temporal_impact, 8))

# Industry impact
industry_impact <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  group_by(industry) %>%
  summarise(
    total_businesses = n(),
    filtered_out = sum(bias_filter),
    kept_for_analysis = total_businesses - filtered_out,
    filter_rate = round(filtered_out / total_businesses * 100, 1),
    .groups = "drop"
  )


print(industry_impact)









# SAVE ENHANCED DATA


output_file <- "data/df_enhanced_with_filters.csv"
write_csv(enhanced_ultra, output_file)




```


##### add street

```{r}



df <- read_csv("data/df_enhanced_with_filters.csv", show_col_types = FALSE)


# Full street types 
street_types_full <- c(
  "Boulevard","Parade","Avenue","Street","Road","Lane","Place","Court","Drive","Walk",
  "Terrace","Highway","Way","Crescent","Square","Mall","Arcade","Circuit","Grove","Close",
  "Esplanade","Promenade","Circus","Quay","Alley","Market","Bridge","Freeway","Causeway",
  "Broadway","Link","Row","Mews","Path","Steps","Rise","Track","Expressway","Bypass"
)

# Abbreviations to expand AFTER we match
abbr_map <- c(
  "\\bSt\\b"  = "Street",
  "\\bRd\\b"  = "Road",
  "\\bAve\\b" = "Avenue",
  "\\bLn\\b"  = "Lane",
  "\\bPde\\b" = "Parade",
  "\\bBlvd\\b"= "Boulevard",
  "\\bBvd\\b" = "Boulevard",
  "\\bPl\\b"  = "Place",
  "\\bCt\\b"  = "Court",
  "\\bDr\\b"  = "Drive",
  "\\bTce\\b" = "Terrace",
  "\\bCres\\b"= "Crescent",
  "\\bEspl\\b"= "Esplanade",
  "\\bHwy\\b" = "Highway",
  "\\bWy\\b"  = "Way",
  "\\bArc\\b" = "Arcade"
)

# Accept either full types OR common abbreviations 
types_alt_with_abbr <- paste(
  c(street_types_full,
    "St","Rd","Ave","Ln","Pde","Blvd","Bvd","Pl","Ct","Dr","Tce","Cres","Espl","Hwy","Wy","Arc"),
  collapse = "|"
)

# Drop common unit/building prefixes at the start
prefix_re <- regex(
  "^(Shop|Unit|Suite|Level|Lvl|Ground|Floor|Tenancy|Kiosk|Building|Bldg|Block|Office|Room|Rm|Apt|Apartment|Podium|Mezzanine)\\s*[A-Za-z0-9\\-/]*\\s*,?\\s*",
  ignore_case = TRUE
)

# Core extractor 
extract_street <- function(address) {
  if (is.na(address) || str_trim(address) == "") return(NA_character_)

  # Normalise + remove unit/building prefixes
  addr <- str_squish(address)
  addr <- str_remove(addr, prefix_re)

  # Pattern: "<Name tokens> <Type|Abbrev>"
  # Works for "Little Latrobe St", "Riverside Quay", "Collins Street", etc.
  pattern <- paste0("([A-Za-z][A-Za-z\\s'\\-]*?)\\s+(", types_alt_with_abbr, ")\\b")

  m <- str_extract(addr, regex(pattern, ignore_case = TRUE))

  # If not found, retry within comma-separated parts (helps with locality suffixes)
  if (is.na(m) && str_detect(addr, ",")) {
    parts <- str_split(addr, ",")[[1]]
    for (p in parts) {
      mm <- str_extract(str_squish(p), regex(pattern, ignore_case = TRUE))
      if (!is.na(mm)) { m <- mm; break }
    }
  }
  if (is.na(m)) return(NA_character_)

  # Remove leading articles OR single-letter tokens (e.g., "A Bourke Street")
  m <- str_remove(m, regex("^(?i)(a|an|the)\\s+"))
  m <- str_remove(m, regex("^([A-Z])\\s+(?=[A-Za-z])"))

  # Expand abbreviations inside the matched phrase
  for (pat in names(abbr_map)) {
    m <- str_replace_all(m, regex(pat, ignore_case = TRUE), abbr_map[[pat]])
  }

  # Title case and tidy
  str_to_title(str_squish(m))
}

# Apply + derive street_type 
df_with_streets <- df %>%
  mutate(
    street_address = map_chr(business_address, extract_street),
    street_type    = str_extract(street_address, paste(street_types_full, collapse = "|"))
  )



# Save outputs 
write_csv(df_with_streets, "data/df_with_streets.csv")





```




```{r}



df <- read_csv("data/df_with_streets.csv", show_col_types = FALSE)

# replacement in street_address
df$street_address[df$street_address == "Alfred Deakin Building Federation Square"] <- "Federation Square"
df$street_address[df$street_address == "Cafe Royal Botanic Gardens Alexandra Avenue"] <- "Alexandra Avenue"
df$street_address[df$street_address == "Montefiore House Exhibition Mews"] <- "Exhibition Mews"
df$street_address[df$street_address == "Restaurant Melbourne Zoological Gardens Elliott Avenue"] <- "Elliott Avenue"
df$street_address[df$street_address == "Restaurant Old Observatory Building Royal Botanic Gardens Birdwood Avenue"] <- "Birdwood Avenue"
df$street_address[df$street_address == "Shed F Queen Victoria Market"] <- "Victoria Market"
df$street_address[df$street_address == "Shed M Queen Victoria Market"] <- "Victoria Market"
df$street_address[df$street_address == "Square Federation Square"] <- "Federation Square"
df$street_address[df$street_address == "Street Kilda Road"] <- "St Kilda Road"
df$street_address[df$street_address == "Street Andrews Place"] <- "St Andrews Place"
df$street_address[df$street_address == "Tea Rooms Royal Botanic Gardens Alexandra Avenue"] <- "Alexandra Avenue"
df$street_address[df$street_address == "Yarra Building Federation Square"] <- "Federation Square"
df$street_address[df$street_address == "Zoological Gardens Elliott Avenue"] <- "Elliott Avenue"

# Remove street_type 
df <- df %>% dplyr::select(-any_of("street_type"))

# Save
write_csv(df, "data/df_with_streets.csv")

# Also save updated unique streets
unique_streets <- df %>%
  filter(!is.na(street_address)) %>%
  distinct(street_address) %>%
  arrange(street_address)

write_csv(unique_streets, "data/unique_streets.csv")




```




#### reliabiluty
```{r}
library(tidyverse)

# Load data
df <- read_csv("data/df_with_streets.csv")



# Get one row per business
business_level <- df %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup()


# Calculator all streets


street_stats_all <- business_level %>%
  filter(
    !is.na(street_address),      # Must have street
    bias_filter == FALSE          # Bias-corrected sample
  ) %>%
  group_by(street_address) %>%
  summarise(
    total_businesses = n(),
    closed_businesses = sum(!str_detect(business_status, "Operating")),
    operating_12plus = sum(str_detect(business_status, "Long-term Success")),
    
    # calculated for ALL streets regardless of sample size
    street_avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    street_median_lifespan = median(lifespan_years, na.rm = TRUE),
    street_success_rate = mean(lifespan_group == ">12 yrs") * 100,
    
    street_avg_seats = mean(num_seats, na.rm = TRUE),
    street_median_seats = median(num_seats, na.rm = TRUE),
    
    # Add reliability indicator based on sample size
    reliability = case_when(
      total_businesses >= 10 ~ "High (10+)",
      total_businesses >= 5 ~ "Medium (5-9)",
      total_businesses >= 3 ~ "Low (3-4)",
      TRUE ~ "Very Low (1-2)"
    ),
    
    .groups = "drop"
  ) %>%
  arrange(desc(street_success_rate))

cat("Total streets with calculated stats:", nrow(street_stats_all), "\n\n")

# Show distribution by reliability
cat("STREET DISTRIBUTION BY SAMPLE SIZE:\n")
reliability_dist <- street_stats_all %>%
  count(reliability) %>%
  mutate(pct = round(100 * n / sum(n), 1))
print(reliability_dist)


# Join to full dadaset


df_final <- df %>%
  left_join(
    street_stats_all %>% 
      dplyr::select(street_address, street_avg_lifespan, street_success_rate, 
             street_median_seats, total_businesses, reliability),
    by = "street_address"
  ) %>%
  mutate(
    street_stats_imputed = is.na(street_address)
  )

# Check coverage
coverage <- df_final %>%
  summarise(
    total_records = n(),
    has_calculated_stats = sum(!is.na(street_avg_lifespan)),
    needs_imputation = sum(is.na(street_avg_lifespan))
  )



if (coverage$needs_imputation > 0) {
  cat("\nImputing for records with no extractable street...\n")
  
  overall_medians <- df_final %>%
    filter(!is.na(street_avg_lifespan)) %>%
    summarise(
      median_lifespan = median(street_avg_lifespan),
      median_success = median(street_success_rate),
      median_seats = median(street_median_seats)
    )
  
  df_final <- df_final %>%
    mutate(
      street_avg_lifespan = coalesce(street_avg_lifespan, overall_medians$median_lifespan),
      street_success_rate = coalesce(street_success_rate, overall_medians$median_success),
      street_median_seats = coalesce(street_median_seats, overall_medians$median_seats),
      total_businesses = coalesce(total_businesses, 0L),
      reliability = coalesce(reliability, "Imputed")
    )
}



# SAVE


write_csv(df_final, "df_with_street_stats_ALL_STREETS.csv")
write_csv(street_stats_all, "street_statistics_ALL_317.csv")


```
















#### Prepare model data

```{r}
library(tidyverse)
library(sf)



# Load source data
df_source <- read_csv("df_with_street_stats_ALL_STREETS.csv")

# Prepare spatial data
business_entry <- df_source %>%
  group_by(business_id) %>%
  arrange(census_year) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  dplyr::select(business_id, entry_year = census_year, industry, lat, lon) %>%
  filter(!is.na(lat), !is.na(lon))

valid_business_ids <- business_entry$business_id

business_sf <- st_as_sf(
  business_entry,
  coords = c("lon", "lat"),
  crs = 4326
) %>% st_transform(28355)

# Calculate competition at 200m
calculate_competition <- function(business_sf, radius_m) {
  
  n <- nrow(business_sf)
  
  competition <- data.frame(
    business_id = business_sf$business_id,
    entry_year = business_sf$entry_year,
    industry = business_sf$industry,
    comp_total = numeric(n),
    comp_same = numeric(n),
    comp_other = numeric(n)
  )
  
  buffers <- st_buffer(business_sf, dist = radius_m)
  pb <- txtProgressBar(max = n, style = 3)
  
  for (i in 1:n) {
    if (i %% 500 == 0) setTxtProgressBar(pb, i)
    
    focal_year <- business_sf$entry_year[i]
    focal_industry <- business_sf$industry[i]
    existing_mask <- business_sf$entry_year <= focal_year
    
    # Total competition
    overlaps_total <- st_intersects(buffers[i,], business_sf[existing_mask,], sparse = FALSE)
    competition$comp_total[i] <- max(0, sum(overlaps_total) - 1)
    
    # Same industry
    same_mask <- existing_mask & business_sf$industry == focal_industry
    overlaps_same <- st_intersects(buffers[i,], business_sf[same_mask,], sparse = FALSE)
    competition$comp_same[i] <- max(0, sum(overlaps_same) - 1)
    
    # Other industry
    other_mask <- existing_mask & business_sf$industry != focal_industry
    overlaps_other <- st_intersects(buffers[i,], business_sf[other_mask,], sparse = FALSE)
    competition$comp_other[i] <- max(0, sum(overlaps_other))
  }
  
  close(pb)
  return(competition)
}

# Calculate 200m competition
comp_200m <- calculate_competition(business_sf, 200)

# Prepare base modeling data
business_level <- df_source %>%
  group_by(business_id) %>%
  arrange(census_year) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(bias_filter == FALSE) %>%
  filter(business_id %in% valid_business_ids)

# Create CLUE area dummies
clue_counts <- business_level %>%
  filter(!is.na(clue_area)) %>%
  count(clue_area, sort = TRUE)

top_clue <- head(clue_counts$clue_area, 10)

# Create street quality indicators
street_perf <- business_level %>%
  filter(!is.na(street_address)) %>%
  group_by(street_address) %>%
  summarise(
    n_businesses = n(),
    avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    success_rate = mean(lifespan_group == ">12 yrs") * 100,
    .groups = "drop"
  ) %>%
  filter(n_businesses >= 5)

success_p80 <- quantile(street_perf$success_rate, 0.80)
success_p20 <- quantile(street_perf$success_rate, 0.20)
lifespan_p80 <- quantile(street_perf$avg_lifespan, 0.80)
lifespan_p20 <- quantile(street_perf$avg_lifespan, 0.20)

best_streets <- street_perf %>%
  filter(success_rate >= success_p80 & avg_lifespan >= lifespan_p80) %>%
  pull(street_address)

worst_streets <- street_perf %>%
  filter(success_rate <= success_p20 & avg_lifespan <= lifespan_p20) %>%
  pull(street_address)

# Build final dataset
df_model <- business_level %>%
  inner_join(comp_200m %>% dplyr::select(business_id, comp_total, comp_same, comp_other), 
             by = "business_id")

df_features <- df_model %>%
  mutate(
    lifespan_numeric = lifespan_years,
    industry_cafe = as.numeric(industry == "Cafes"),
    log_seats = log1p(num_seats),
    outdoor_seating = as.numeric(str_detect(seating_type, "Outdoor")),
    mixed_seating = as.numeric(str_detect(seating_type, "Mixed")),
    same_ind = comp_same,
    other_ind = comp_other,
    on_best_street = as.numeric(street_address %in% best_streets),
    on_worst_street = as.numeric(street_address %in% worst_streets),
    property_competition = businesses_at_property - 1
  )

# Add CLUE area dummies
for (area in top_clue) {
  col_name <- paste0("clue_", str_replace_all(str_to_lower(area), "[^a-z0-9]", "_"))
  df_features[[col_name]] <- as.numeric(df_features$clue_area == area)
}

# Select final features
final_features <- c(
  "lifespan_numeric","business_id",
  "industry_cafe", "log_seats", "outdoor_seating", "mixed_seating",
  "same_ind", "other_ind",
  "on_best_street", "on_worst_street",
  paste0("clue_", str_replace_all(str_to_lower(top_clue), "[^a-z0-9]", "_")),
  "property_competition"
)

df_final <- df_features %>%
  dplyr::select(all_of(final_features))

# Save
write_csv(df_final, "df_clean_200m_shinny.csv")

# Summary
summary_stats <- tibble(
  metric = c("Total rows", "Features", "Mean same_ind", "Mean other_ind", "NAs present"),
  value = c(
    nrow(df_final),
    ncol(df_final),
    round(mean(df_final$same_ind), 1),
    round(mean(df_final$other_ind), 1),
    sum(is.na(df_final))
  )
)

print(summary_stats)


```











```{r}




# Load the Ultra final data
df_ultra <- read_csv("data/df_FINAL_2industries.csv")

cat("Loaded", nrow(df_ultra), "records from Ultra data\n")
cat("Unique businesses:", n_distinct(df_ultra$business_id), "\n")


# CALCULATE BUSINESS OPERATING STATUS


business_status <- df_ultra %>%
  group_by(business_id) %>%
  summarise(
    business_start_year = min(census_year),
    business_end_year = max(census_year),
    still_operating = (business_end_year == 2023),  # 2023 is the latest year in dataset
    lifespan_group = first(lifespan_group),
    .groups = "drop"
  )



# ADD BIAS FILTER LOGIC


enhanced_ultra <- df_ultra %>%
  left_join(business_status %>% select(business_id, business_start_year, business_end_year, still_operating), 
            by = "business_id") %>%
  mutate(
    # BIAS FILTER LOGIC:
    # TRUE = exclude from bias-corrected analysis (but keep for temporal analysis)
    # FALSE = include in all analysis
    bias_filter = case_when(
      # Keep ALL >12 year businesses (whether operating or closed)
      lifespan_group == ">12 yrs" ~ FALSE,
      
      # For shorter lifespan groups: only keep closed businesses
      lifespan_group != ">12 yrs" & !still_operating ~ FALSE,  # Closed = keep (true lifespan)
      lifespan_group != ">12 yrs" & still_operating ~ TRUE,    # Still operating = filter out (biased lifespan)
      
      TRUE ~ FALSE
    ),
    
    # Add descriptive status for transparency
    business_status = case_when(
      lifespan_group == ">12 yrs" & still_operating ~ "Long-term Success (Operating)",
      lifespan_group == ">12 yrs" & !still_operating ~ "Long-term Success (Closed)",
      !still_operating ~ "Closed Business",
      still_operating ~ "Still Operating (Biased Lifespan)",
      TRUE ~ "Unknown"
    ),
    
    # Add analysis recommendation
    analysis_usage = case_when(
      bias_filter == FALSE ~ "Use in All Analysis",
      bias_filter == TRUE ~ "Temporal Only (Exclude from Lifespan Analysis)",
      TRUE ~ "Unknown"
    )
  ) %>%
  # Clean up temporary columns
  select(-business_start_year, -business_end_year, -still_operating) %>%
  # Arrange for consistency
  arrange(census_year, property_id, business_id)


# QUALITY ANALYSIS & VALIDATION




# Overall filtering impact
total_records <- nrow(enhanced_ultra)
filtered_records <- sum(enhanced_ultra$bias_filter)
kept_records <- total_records - filtered_records



# Business-level analysis
business_analysis <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  count(lifespan_group, business_status) %>%
  pivot_wider(names_from = business_status, values_from = n, values_fill = 0)


print(business_analysis)

# Lifespan group impact
lifespan_impact <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  group_by(lifespan_group) %>%
  summarise(
    total_businesses = n(),
    filtered_out = sum(bias_filter),
    kept_for_analysis = total_businesses - filtered_out,
    filter_rate = round(filtered_out / total_businesses * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(lifespan_group)


print(lifespan_impact)

# Area impact analysis
area_impact <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  group_by(clue_area) %>%
  summarise(
    total_businesses = n(),
    filtered_out = sum(bias_filter),
    kept_for_analysis = total_businesses - filtered_out,
    filter_rate = round(filtered_out / total_businesses * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(total_businesses))


print(head(area_impact, 10))

# Temporal impact by year
temporal_impact <- enhanced_ultra %>%
  group_by(census_year) %>%
  summarise(
    total_records = n(),
    filtered_records = sum(bias_filter),
    kept_records = total_records - filtered_records,
    filter_rate = round(filtered_records / total_records * 100, 1),
    .groups = "drop"
  )


print(tail(temporal_impact, 8))

# Industry impact
industry_impact <- enhanced_ultra %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  group_by(industry) %>%
  summarise(
    total_businesses = n(),
    filtered_out = sum(bias_filter),
    kept_for_analysis = total_businesses - filtered_out,
    filter_rate = round(filtered_out / total_businesses * 100, 1),
    .groups = "drop"
  )


print(industry_impact)


# VERIFY COLUMN NAMES MATCH



print(names(enhanced_ultra))

# Expected columns based on enhanced_with_filters structure
expected_columns <- c(
  "property_id", "business_id", "census_year", "industry", "clue_area",
  "trading_name", "business_address", "building_address", "seating_type",
  "num_seats", "lifespan_years", "lifespan_group", "lon", "lat",
  "businesses_at_property", "bias_filter", "business_status", "analysis_usage"
)


print(expected_columns)




# USAGE EXAMPLES FOR ANALYSIS



# Sample data to show the structure

sample_data <- enhanced_ultra %>%
  select(business_id, census_year, trading_name, clue_area, lifespan_group, 
         bias_filter, business_status, analysis_usage) %>%
  head(10)
print(sample_data)


# SAVE ENHANCED DATA


output_file <- "df_ULTRA_enhanced_with_filters.csv"
write_csv(enhanced_ultra, output_file)


```


























```{r}


library(tidyverse)
library(sf)
library(lubridate)


# load data



# Base dataset (current Shiny-ready data)
df_base <- read_csv("df_ULTRA_enhanced_with_filters.csv")

# Street statistics (from previous processing)
df_streets <- read_csv("df_with_street_stats_ALL_STREETS.csv")

# Modeling data with 200m competition (extract competition metrics only)
df_model <- read_csv("df_clean_200m_shinny.csv")




# Add foot trsffic data




# Extract unique street info (one row per business)
street_lookup <- df_streets %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  dplyr::select(
    business_id,
    street_address,
    street_avg_lifespan,
    street_success_rate,
    street_median_seats,
    total_businesses_on_street = total_businesses,
    street_reliability = reliability
  )

# Merge with base
df_enhanced <- df_base %>%
  left_join(street_lookup, by = "business_id")

cat("✓ Added street data -", 
    sum(!is.na(df_enhanced$street_address)), "records have street addresses\n\n")


# work out 50m competition




# Get business entry points (first appearance year for each business)
business_entries <- df_base %>%
  group_by(business_id) %>%
  arrange(census_year) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  dplyr::select(business_id, entry_year = census_year, industry, lat, lon) %>%
  filter(!is.na(lat), !is.na(lon))

# Convert to spatial object
business_sf <- st_as_sf(
  business_entries,
  coords = c("lon", "lat"),
  crs = 4326  # WGS84
) %>% 
  st_transform(28355)  # Transform to projected CRS (meters)

# Calculate 50m competition
radius_50m <- 50
n <- nrow(business_sf)

competition_50m <- data.frame(
  business_id = business_sf$business_id,
  entry_year = business_sf$entry_year,
  industry = business_sf$industry,
  competitors_50m = numeric(n),
  same_industry_50m = numeric(n),
  other_industry_50m = numeric(n)
)

# Create 50m buffers
buffers_50m <- st_buffer(business_sf, dist = radius_50m)

# Progress bar
pb <- txtProgressBar(max = n, style = 3)

for (i in 1:n) {
  if (i %% 500 == 0) {
    setTxtProgressBar(pb, i)
    cat(sprintf(" [%d/%d - %.1f%%]\n", i, n, i/n*100))
  }
  
  focal_year <- business_sf$entry_year[i]
  focal_industry <- business_sf$industry[i]
  
  # Find businesses that existed at entry year
  existing_mask <- business_sf$entry_year <= focal_year
  
  # Total competitors (exclude self)
  overlaps_total <- st_intersects(buffers_50m[i,], business_sf[existing_mask,], sparse = FALSE)
  competition_50m$competitors_50m[i] <- max(0, sum(overlaps_total) - 1)
  
  # Same industry competitors
  same_mask <- existing_mask & business_sf$industry == focal_industry
  overlaps_same <- st_intersects(buffers_50m[i,], business_sf[same_mask,], sparse = FALSE)
  competition_50m$same_industry_50m[i] <- max(0, sum(overlaps_same) - 1)
  
  # Other industry competitors
  other_mask <- existing_mask & business_sf$industry != focal_industry
  overlaps_other <- st_intersects(buffers_50m[i,], business_sf[other_mask,], sparse = FALSE)
  competition_50m$other_industry_50m[i] <- max(0, sum(overlaps_other))
}

close(pb)






# Add 200m competition




# Extract 200m competition (from modeling dataset)
# Note: modeling data has one row per business (entry year only)
competition_200m <- df_model %>%
  dplyr::select(
    business_id,
    same_industry_200m = same_ind,
    other_industry_200m = other_ind
  ) %>%
  mutate(competitors_200m = same_industry_200m + other_industry_200m)




# Merge data

# Merge 50m competition (matches by business_id for each census year)
df_enhanced <- df_enhanced %>%
  left_join(
    competition_50m %>% dplyr::select(business_id, competitors_50m, same_industry_50m, other_industry_50m),
    by = "business_id"
  )

# Merge 200m competition
df_enhanced <- df_enhanced %>%
  left_join(competition_200m, by = "business_id")




# Add features for shinny app




df_enhanced <- df_enhanced %>%
  mutate(
    # ---- Competition Categories (for filtering) ----
    competition_50m_category = case_when(
      is.na(competitors_50m) ~ "Unknown",
      competitors_50m == 0 ~ "Isolated (0)",
      competitors_50m <= 2 ~ "Very Low (1-2)",
      competitors_50m <= 5 ~ "Low (3-5)",
      competitors_50m <= 10 ~ "Medium (6-10)",
      competitors_50m <= 20 ~ "High (11-20)",
      TRUE ~ "Very High (>20)"
    ),
    
    competition_200m_category = case_when(
      is.na(competitors_200m) ~ "Unknown",
      competitors_200m == 0 ~ "Isolated (0)",
      competitors_200m <= 5 ~ "Very Low (1-5)",
      competitors_200m <= 15 ~ "Low (6-15)",
      competitors_200m <= 30 ~ "Medium (16-30)",
      competitors_200m <= 50 ~ "High (31-50)",
      TRUE ~ "Very High (>50)"
    ),
    
    #  Competition Flags (binary) 
    high_competition_50m = as.numeric(competitors_50m >= 10),
    high_competition_200m = as.numeric(competitors_200m >= 30),
    isolated_location = as.numeric(competitors_50m == 0 | competitors_200m <= 2),
    
    #  Street Quality Indicators 
    # Calculate percentiles for street success rate
    street_quality = case_when(
      is.na(street_success_rate) ~ "Unknown",
      street_success_rate >= quantile(street_success_rate, 0.8, na.rm = TRUE) ~ "Excellent (Top 20%)",
      street_success_rate >= quantile(street_success_rate, 0.6, na.rm = TRUE) ~ "Good (60-80%)",
      street_success_rate >= quantile(street_success_rate, 0.4, na.rm = TRUE) ~ "Average (40-60%)",
      street_success_rate >= quantile(street_success_rate, 0.2, na.rm = TRUE) ~ "Below Average (20-40%)",
      TRUE ~ "Poor (Bottom 20%)"
    ),
    
    on_prime_street = as.numeric(
      !is.na(street_success_rate) & 
      street_success_rate >= quantile(street_success_rate, 0.8, na.rm = TRUE)
    ),
    
    on_poor_street = as.numeric(
      !is.na(street_success_rate) & 
      street_success_rate <= quantile(street_success_rate, 0.2, na.rm = TRUE)
    ),
    
    # Industry Competition Ratios 
    industry_ratio_50m = case_when(
      is.na(competitors_50m) | competitors_50m == 0 ~ NA_real_,
      TRUE ~ same_industry_50m / competitors_50m
    ),
    
    industry_ratio_200m = case_when(
      is.na(competitors_200m) | competitors_200m == 0 ~ NA_real_,
      TRUE ~ same_industry_200m / competitors_200m
    ),
    
    # Size Categories (for analysis) 
    size_category = case_when(
      num_seats <= 30 ~ "Small (≤30)",
      num_seats <= 60 ~ "Medium (31-60)",
      num_seats <= 100 ~ "Large (61-100)",
      TRUE ~ "Very Large (>100)"
    ),
    
    # CBD Distance (approximate) 
    # CBD center: 144.9631°E, 37.8136°S
    dist_from_cbd_km = sqrt(
      (lon - 144.9631)^2 + (lat - (-37.8136))^2
    ) * 111,  # Rough conversion to km
    
    # Current Status (for 2023 snapshot) 
    is_current_business = as.numeric(census_year == 2023),
    
    #  Risk Flags (based on modeling insights) 
    high_risk_location = as.numeric(
      (on_poor_street == 1) |
      (clue_area == "Docklands") |
      (other_industry_200m >= 25)
    ),
    
    favorable_location = as.numeric(
      (on_prime_street == 1) |
      (clue_area %in% c("Melbourne (CBD)", "Carlton")) |
      (other_industry_200m <= 15)
    )
  )




# Work out street level data




# Calculate current (2023) business counts per street
current_street_stats <- df_enhanced %>%
  filter(census_year == 2023, !is.na(street_address)) %>%
  group_by(street_address, industry) %>%
  summarise(
    current_businesses_2023 = n(),
    avg_seats_current = mean(num_seats, na.rm = TRUE),
    median_competitors_50m = median(competitors_50m, na.rm = TRUE),
    median_competitors_200m = median(competitors_200m, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate historical performance per street
historical_street_stats <- df_enhanced %>%
  filter(bias_filter == FALSE) %>%  # Use only unbiased sample
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(street_address)) %>%
  group_by(street_address, industry) %>%
  summarise(
    total_businesses_historical = n(),
    avg_lifespan_historical = mean(lifespan_years, na.rm = TRUE),
    median_lifespan_historical = median(lifespan_years, na.rm = TRUE),
    success_rate_12yrs = mean(lifespan_group == ">12 yrs") * 100,
    avg_seats_historical = mean(num_seats, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(total_businesses_historical >= 3)  # Only streets with 3+ historical businesses

# Combine street statistics
street_analysis_data <- historical_street_stats %>%
  left_join(
    current_street_stats,
    by = c("street_address", "industry")
  ) %>%
  mutate(
    # Add CLUE area for comparison (use most common clue_area for this street)
    clue_area = df_enhanced %>%
      filter(street_address == .$street_address[1]) %>%
      count(clue_area, sort = TRUE) %>%
      slice_head(n = 1) %>%
      pull(clue_area)
  )

# Add CLUE area benchmarks
clue_benchmarks <- df_enhanced %>%
  filter(bias_filter == FALSE) %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  group_by(clue_area, industry) %>%
  summarise(
    clue_avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    clue_success_rate = mean(lifespan_group == ">12 yrs") * 100,
    .groups = "drop"
  )

street_analysis_data <- street_analysis_data %>%
  left_join(clue_benchmarks, by = c("industry")) %>%
  mutate(
    vs_area_benchmark = avg_lifespan_historical - clue_avg_lifespan,
    better_than_area = vs_area_benchmark > 0
  )










# Save

# Main enhanced dataset (for Shiny app)
write_csv(df_enhanced, "df_ULTRA_enhanced_with_competition.csv")


# Street analysis lookup (for Street Analysis Tool)
write_csv(street_analysis_data, "street_analysis_lookup.csv")


# Competition metrics summary (for documentation)
competition_summary <- df_enhanced %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  filter(!is.na(competitors_200m)) %>%
  summarise(
    total_businesses = n(),
    mean_comp_50m = round(mean(competitors_50m, na.rm = TRUE), 1),
    median_comp_50m = median(competitors_50m, na.rm = TRUE),
    mean_comp_200m = round(mean(competitors_200m, na.rm = TRUE), 1),
    median_comp_200m = median(competitors_200m, na.rm = TRUE),
    pct_high_comp = round(mean(high_competition_200m, na.rm = TRUE) * 100, 1),
    pct_isolated = round(mean(isolated_location, na.rm = TRUE) * 100, 1)
  )

write_csv(competition_summary, "competition_metrics_summary.csv")



```





```{r}


library(tidyverse)
library(sf)



# Revise 200m competition



# Load current data
df_current <- read_csv("df_ULTRA_enhanced_with_competition.csv", show_col_types = FALSE)

# Remove old incomplete 200m columns
df_fixed <- df_current %>%
  dplyr::select(-competitors_200m, -same_industry_200m, -other_industry_200m, 
         -competition_200m_category, -high_competition_200m, -industry_ratio_200m)



# Get business entry points (ALL businesses with coordinates)
business_entries <- df_fixed %>%
  group_by(business_id) %>%
  arrange(census_year) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  dplyr::select(business_id, entry_year = census_year, industry, lat, lon) %>%
  filter(!is.na(lat), !is.na(lon))



# Convert to spatial object
business_sf <- st_as_sf(
  business_entries,
  coords = c("lon", "lat"),
  crs = 4326
) %>% 
  st_transform(28355)

# Calculate 200m competition
radius_200m <- 200
n <- nrow(business_sf)



competition_200m_new <- data.frame(
  business_id = business_sf$business_id,
  entry_year = business_sf$entry_year,
  industry = business_sf$industry,
  competitors_200m = numeric(n),
  same_industry_200m = numeric(n),
  other_industry_200m = numeric(n)
)

# Create 200m buffers
buffers_200m <- st_buffer(business_sf, dist = radius_200m)

# Progress tracking
pb <- txtProgressBar(max = n, style = 3)
start_time <- Sys.time()

for (i in 1:n) {
  if (i %% 500 == 0) {
    elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))
    pct_complete <- i/n
    est_total <- elapsed / pct_complete
    est_remaining <- est_total - elapsed
    
    setTxtProgressBar(pb, i)
    cat(sprintf("\n[%d/%d - %.1f%%] Elapsed: %.1f min | Est. remaining: %.1f min\n", 
                i, n, pct_complete*100, elapsed, est_remaining))
  }
  
  focal_year <- business_sf$entry_year[i]
  focal_industry <- business_sf$industry[i]
  
  # Find businesses that existed at entry year
  existing_mask <- business_sf$entry_year <= focal_year
  
  # Total competitors (exclude self)
  overlaps_total <- st_intersects(buffers_200m[i,], business_sf[existing_mask,], sparse = FALSE)
  competition_200m_new$competitors_200m[i] <- max(0, sum(overlaps_total) - 1)
  
  # Same industry competitors
  same_mask <- existing_mask & business_sf$industry == focal_industry
  overlaps_same <- st_intersects(buffers_200m[i,], business_sf[same_mask,], sparse = FALSE)
  competition_200m_new$same_industry_200m[i] <- max(0, sum(overlaps_same) - 1)
  
  # Other industry competitors
  other_mask <- existing_mask & business_sf$industry != focal_industry
  overlaps_other <- st_intersects(buffers_200m[i,], business_sf[other_mask,], sparse = FALSE)
  competition_200m_new$other_industry_200m[i] <- max(0, sum(overlaps_other))
}

close(pb)

total_time <- as.numeric(difftime(Sys.time(), start_time, units = "mins"))
cat(sprintf("\n✓ 200m competition calculated in %.1f minutes\n\n", total_time))




# Merge data



# Merge with main dataset
df_fixed <- df_fixed %>%
  left_join(
    competition_200m_new %>% dplyr::select(business_id, competitors_200m, same_industry_200m, other_industry_200m),
    by = "business_id"
  )



# Recalculate 200m derived features
df_fixed <- df_fixed %>%
  mutate(
    # 200m competition category
    competition_200m_category = case_when(
      is.na(competitors_200m) ~ "Unknown",
      competitors_200m == 0 ~ "Isolated (0)",
      competitors_200m <= 5 ~ "Very Low (1-5)",
      competitors_200m <= 15 ~ "Low (6-15)",
      competitors_200m <= 30 ~ "Medium (16-30)",
      competitors_200m <= 50 ~ "High (31-50)",
      TRUE ~ "Very High (>50)"
    ),
    
    # High competition flag
    high_competition_200m = as.numeric(competitors_200m >= 30),
    
    # Industry ratio
    industry_ratio_200m = case_when(
      is.na(competitors_200m) | competitors_200m == 0 ~ NA_real_,
      TRUE ~ same_industry_200m / competitors_200m
    ),
    
    # Update high risk location (now with complete 200m data)
    high_risk_location = as.numeric(
      (on_poor_street == 1) |
      (clue_area == "Docklands") |
      (!is.na(other_industry_200m) & other_industry_200m >= 25)
    ),
    
    # Update favorable location
    favorable_location = as.numeric(
      (on_prime_street == 1) |
      (clue_area %in% c("Melbourne (CBD)", "Carlton")) |
      (!is.na(other_industry_200m) & other_industry_200m <= 15)
    )
  )



# Validation check
cat("Validation - 200m data coverage:\n")
business_check <- df_fixed %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  summarise(
    total = n(),
    has_50m = sum(!is.na(competitors_50m)),
    has_200m = sum(!is.na(competitors_200m)),
    has_both = sum(!is.na(competitors_50m) & !is.na(competitors_200m))
  )

print(business_check)





# FIX STREET ANALYSIS LOOKUP




# Load and fix street lookup
street_lookup_old <- read_csv("street_analysis_lookup.csv", show_col_types = FALSE)





# Fix: Use clue_area.x (from historical data join) as primary
street_lookup_fixed <- street_lookup_old %>%
  mutate(
    # Prefer clue_area.x, fallback to clue_area.y if missing
    clue_area = coalesce(clue_area.x, clue_area.y)
  ) %>%
  dplyr::select(-clue_area.x, -clue_area.y) %>%
  # Reorder columns logically
  dplyr::select(
    street_address, 
    industry,
    clue_area,
    # Historical metrics
    total_businesses_historical,
    avg_lifespan_historical,
    median_lifespan_historical,
    success_rate_12yrs,
    avg_seats_historical,
    # Current metrics
    current_businesses_2023,
    avg_seats_current,
    median_competitors_50m,
    median_competitors_200m,
    # Benchmark comparison
    clue_avg_lifespan,
    clue_success_rate,
    vs_area_benchmark,
    better_than_area
  )



# Add additional street-level competition stats using new 200m data
street_competition_stats <- df_fixed %>%
  filter(census_year == 2023, !is.na(street_address)) %>%
  group_by(street_address, industry) %>%
  summarise(
    mean_competitors_50m = round(mean(competitors_50m, na.rm = TRUE), 1),
    mean_competitors_200m = round(mean(competitors_200m, na.rm = TRUE), 1),
    pct_high_competition = round(mean(high_competition_200m, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

street_lookup_fixed <- street_lookup_fixed %>%
  left_join(street_competition_stats, by = c("street_address", "industry"))


# Save



# Save fixed main dataset
write_csv(df_fixed, "df_ULTRA_enhanced_with_competition_FIXED.csv")

# Save fixed street lookup
write_csv(street_lookup_fixed, "street_analysis_lookup_FIXED.csv")






```




#### Add foot traffic data

```{r}


library(tidyverse)
library(sf)




# Load business data
df_business <- read_csv("df_ULTRA_enhanced_with_competition_FIXED.csv")

# Load foot traffic data
traffic_overview <- read_csv("overview_sensor_traffic.csv")
traffic_yearly <- read_csv("yearly_sensor_traffic.csv")



# Find Nearest Sensor for Each Business

# Convert to spatial objects
business_sf <- df_business %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(28355)  # Melbourne projection

sensors_sf <- traffic_overview %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(28355)



# Find nearest sensor within 200m (sensors beyond this are not representative)
nearest_sensor <- st_nearest_feature(business_sf, sensors_sf)
distances <- st_distance(business_sf, sensors_sf[nearest_sensor,], by_element = TRUE)

# Add sensor info to business data
df_with_traffic <- df_business %>%
  filter(!is.na(lat), !is.na(lon)) %>%
  mutate(
    row_id = row_number(),
    nearest_sensor_id = traffic_overview$Location_ID[nearest_sensor],
    nearest_sensor_name = traffic_overview$Sensor_Name[nearest_sensor],
    sensor_distance_m = as.numeric(distances),
    avg_daily_foottraffic = traffic_overview$avg_daily_overall[nearest_sensor],
    foottraffic_years_observed = traffic_overview$years_observed[nearest_sensor]
  ) %>%
  # Only keep sensor data if within 200m
  mutate(
    avg_daily_foottraffic = ifelse(sensor_distance_m <= 200, avg_daily_foottraffic, NA),
    nearest_sensor_name = ifelse(sensor_distance_m <= 200, nearest_sensor_name, NA)
  )

# Merge back with businesses without coordinates
df_final <- df_business %>%
  left_join(
    df_with_traffic %>% 
      dplyr::select(property_id, business_id, census_year, 
             nearest_sensor_id, nearest_sensor_name, sensor_distance_m, 
             avg_daily_foottraffic, foottraffic_years_observed),
    by = c("property_id", "business_id", "census_year")
  )


# CREATE FOOT TRAFFIC CATEGORIES


df_final <- df_final %>%
  mutate(
    # Categorize foot traffic levels
    foottraffic_category = case_when(
      is.na(avg_daily_foottraffic) ~ "Unknown",
      avg_daily_foottraffic < 1000 ~ "Very Low (<1k/day)",
      avg_daily_foottraffic < 3000 ~ "Low (1-3k/day)",
      avg_daily_foottraffic < 5000 ~ "Medium (3-5k/day)",
      avg_daily_foottraffic < 8000 ~ "High (5-8k/day)",
      TRUE ~ "Very High (>8k/day)"
    ),
    
    # Create traffic density score (per 1000 people)
    traffic_density_k = avg_daily_foottraffic / 1000,
    
    # Flag high-traffic locations (top 25%)
    high_traffic_location = as.numeric(
      !is.na(avg_daily_foottraffic) & 
      avg_daily_foottraffic >= quantile(avg_daily_foottraffic, 0.75, na.rm = TRUE)
    ),
    
    # Flag low-traffic locations (bottom 25%)
    low_traffic_location = as.numeric(
      !is.na(avg_daily_foottraffic) & 
      avg_daily_foottraffic <= quantile(avg_daily_foottraffic, 0.25, na.rm = TRUE)
    )
  )


# Analysis of foot traffic




# Coverage statistics
coverage_stats <- df_final %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  summarise(
    total_businesses = n(),
    with_traffic_data = sum(!is.na(avg_daily_foottraffic)),
    pct_coverage = round(sum(!is.na(avg_daily_foottraffic)) / n() * 100, 1)
  )

cat("Coverage Statistics:\n")
print(coverage_stats)
cat("\n")

# Traffic vs Survival Analysis
traffic_survival <- df_final %>%
  filter(!bias_filter, !is.na(avg_daily_foottraffic)) %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(
    traffic_bin = cut(avg_daily_foottraffic, 
                     breaks = c(0, 2000, 4000, 6000, 8000, 20000),
                     labels = c("<2k", "2-4k", "4-6k", "6-8k", ">8k"))
  ) %>%
  group_by(traffic_bin) %>%
  summarise(
    n_businesses = n(),
    avg_lifespan = round(mean(lifespan_years, na.rm = TRUE), 1),
    median_lifespan = median(lifespan_years, na.rm = TRUE),
    success_rate = round(mean(lifespan_group == ">12 yrs") * 100, 1),
    avg_competition = round(mean(competitors_200m, na.rm = TRUE), 0),
    .groups = "drop"
  )



# Golden Opportunities: High traffic + Low competition
golden_spots <- df_final %>%
  filter(
    census_year == 2023,
    !is.na(avg_daily_foottraffic),
    !is.na(competitors_200m)
  ) %>%
  mutate(
    opportunity_score = (avg_daily_foottraffic / 1000) / (competitors_200m + 1)
  ) %>%
  group_by(street_address) %>%
  summarise(
    n_current = n(),
    avg_foottraffic = round(mean(avg_daily_foottraffic), 0),
    avg_competition = round(mean(competitors_200m), 0),
    avg_opportunity_score = round(mean(opportunity_score), 2),
    clue_area = first(clue_area),
    .groups = "drop"
  ) %>%
  filter(!is.na(street_address), n_current >= 1) %>%
  arrange(desc(avg_opportunity_score)) %>%
  head(20)




# UPDATE street level analysis


street_traffic <- df_final %>%
  filter(census_year == 2023, !is.na(street_address), !is.na(avg_daily_foottraffic)) %>%
  group_by(street_address, industry) %>%
  summarise(
    avg_foottraffic = round(mean(avg_daily_foottraffic), 0),
    median_foottraffic = median(avg_daily_foottraffic),
    .groups = "drop"
  )

# Load and update street lookup
street_lookup <- read_csv("street_analysis_lookup_FIXED.csv")

street_lookup_enhanced <- street_lookup %>%
  left_join(street_traffic, by = c("street_address", "industry"))


# Save
write_csv(df_final, "df_ULTRA_enhanced_with_traffic.csv")
write_csv(street_lookup_enhanced, "street_analysis_lookup_with_traffic.csv")
write_csv(golden_spots, "golden_opportunity_streets.csv")




# CORRELATION ANALYSIS




# Correlation: Foot traffic vs Lifespan
business_level <- df_final %>%
  filter(!bias_filter, !is.na(avg_daily_foottraffic)) %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup()

if (nrow(business_level) > 0) {
  cor_traffic_lifespan <- cor(business_level$avg_daily_foottraffic, 
                               business_level$lifespan_years, 
                               use = "complete.obs")
  
  cor_traffic_success <- cor(business_level$avg_daily_foottraffic, 
                             as.numeric(business_level$lifespan_group == ">12 yrs"), 
                             use = "complete.obs")
  
}

# Sweet spot analysis
sweet_spot <- traffic_survival %>%
  filter(!is.na(success_rate)) %>%
  arrange(desc(success_rate)) %>%
  slice_head(n = 1)


```



```{r}


# 1) Load the attached canonical file
df <- readr::read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)

# 2) Compute first/last seen years per business_id
meta <- df %>%
  group_by(business_id) %>%
  summarise(
    first_seen_year = min(census_year, na.rm = TRUE),
    last_seen_year  = max(census_year, na.rm = TRUE),
    # observed lifespan within the data window (kept separate from your existing column)
    lifespan_years_obs = last_seen_year - first_seen_year + 1,
    .groups = "drop"
  )

# 3) Join meta back and create filters
df_with_flags <- df %>%
  left_join(meta, by = "business_id") %>%
  mutate(
    # Veteran if observed lifespan > 12 *or* your existing group marks it as >12 yrs
    is_veteran_obs = lifespan_years_obs > 12 |
                     (!is.na(lifespan_group) & lifespan_group == ">12 yrs"),

    # NEW: start-year filter to remove left-censoring bias
    # Keep if first seen after 2002 OR is a clear veteran
    start_year_filter = first_seen_year > 2002 | is_veteran_obs,

    # COMPOSITE: reuse your existing bias_filter (handles right-censoring)
    # and add the new start-year rule for lifecycle/model analyses
    include_lifecycle_biasfree = (!bias_filter) & start_year_filter
  )

# 4) (Optional) Quick diagnostics

print(df_with_flags %>% distinct(business_id, start_year_filter) %>% count(start_year_filter))


print(df_with_flags %>% distinct(business_id, include_lifecycle_biasfree) %>% count(include_lifecycle_biasfree))


print(
  df_with_flags %>%
    distinct(business_id, first_seen_year, is_veteran_obs) %>%
    filter(first_seen_year == 2002) %>%
    count(is_veteran_obs)
)

# 5) Save updated dataset (non-destructive new file)
readr::write_csv(df_with_flags, "df_ULTRA_enhanced_with_traffic.csv")


# 6) (Optional) One-row-per-business view for analyses that need a single row per firm at first observation
df_firstrow_with_flags <- df_with_flags %>%
  arrange(business_id, census_year) %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup()

readr::write_csv(df_firstrow_with_flags, "df_firstrow_with_filters.csv")


```


```{r}
library(dplyr)
library(readr)

# If your data frame is already in memory as `df`, skip the read_csv line.
df <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)

# ---- build a business-level table (first observation per business) ----
biz_first <- df %>%
  arrange(business_id, census_year) %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup()

# ---- apply BOTH filters to define the valid population for street stats ----
biz_valid <- biz_first %>%
  filter(bias_filter == FALSE, start_year_filter == TRUE)

# ---- recompute street-level rollups (same names as before) ----
street_rollups <- biz_valid %>%
  group_by(street_address) %>%
   summarise(
    n_businesses        = n(),  # Now truly one row per business
    n_success           = sum(lifespan_group == ">12 yrs", na.rm = TRUE),
    street_avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    street_success_rate = 100 * n_success / n_businesses,
    street_median_seats = mean(num_seats, na.rm = TRUE),
    .groups = "drop"
  )
# ---- overwrite the three columns (names unchanged) and keep everything else ----
df_updated <- df %>%
  select(-any_of(c("street_avg_lifespan", "street_success_rate", "street_median_seats"))) %>%
  left_join(street_rollups, by = "street_address")

# optional: save
write_csv(df_updated, "df_ULTRA_enhanced_with_traffic.csv")


```





#### EDA



```{r}
library(readr)
library(dplyr)
library(ggplot2)

# Load data
df <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)%>%
  filter(start_year_filter == TRUE)

# Prepare yearly counts (unique businesses) for the two industries
industry_year_counts <- df %>%
  filter(industry %in% c("Cafes", "Takeaway")) %>%
  mutate(industry_plot = recode(industry, "Cafes" = "Restaurant", "Takeaway" = "Takeaway")) %>%
  group_by(census_year, industry_plot) %>%
  summarise(Count = n_distinct(business_id), .groups = "drop") %>%
  arrange(census_year, industry_plot)

# Plot: single panel, two lines, dots at each year
ggplot(industry_year_counts, aes(x = census_year, y = Count, color = industry_plot)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    x = "Census Year",
    y = "Business Count",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")

```


```{r}
library(readr)
library(dplyr)
library(ggplot2)

df <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)

# Yearly business counts
growth_data <- df %>%
  filter(industry %in% c("Cafes", "Takeaway")) %>%
  mutate(industry_plot = recode(industry, "Cafes" = "Restaurant", "Takeaway" = "Takeaway")) %>%
  group_by(census_year, industry_plot) %>%
  summarise(Count = n_distinct(business_id), .groups = "drop") %>%
  arrange(industry_plot, census_year) %>%
  group_by(industry_plot) %>%
  mutate(GrowthRate = (Count - lag(Count)) / lag(Count) * 100) %>%
  ungroup()

# Plot: year-to-year growth (%)
p_trand <- ggplot(growth_data, aes(x = census_year, y = GrowthRate, color = industry_plot)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Business Growth Trends Over Time",
    subtitle = "Annual growth patterns across hospitality industries",
    x = "Census Year",
    y = "Growth Rate (%)",
    color = "Industry"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )

p_trand 

```









```{r}
library(readr)
library(dplyr)
library(ggplot2)

# Load data
df <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)

# Define industries of interest
selected_industries <- c("Cafes", "Takeaway")

# Compute average seating per business (ignore seating_type differences)
avg_seating_trend <- df %>%
  filter(industry %in% selected_industries) %>%
  mutate(industry_plot = recode(industry,
                                "Cafes" = "Restaurant",
                                "Takeaway" = "Takeaway")) %>%
  # collapse duplicates within each business per year
  group_by(business_id, census_year, industry_plot) %>%
  summarise(total_seats = max(num_seats, na.rm = TRUE), .groups = "drop") %>%
  # aggregate per year and industry
  group_by(census_year, industry_plot) %>%
  summarise(
    total_seats = sum(total_seats, na.rm = TRUE),
    total_businesses = n_distinct(business_id),
    avg_seats_per_business = total_seats / total_businesses,
    .groups = "drop"
  )

# ---- Plot ----
ggplot(avg_seating_trend,
       aes(x = census_year, y = avg_seats_per_business, color = industry_plot)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.2) +
  labs(
    title = "Average Seating per Business Over Time",
    subtitle = "Comparison between Restaurant (Cafes) and Takeaway industries",
    x = "Census Year",
    y = "Average Number of Seats per Business",
    color = "Industry"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")



```







```{r}


# 1) Load street data (time series by census_year)
df <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)%>%
  filter(start_year_filter == TRUE)

# 2) Collapse to ONE ROW PER BUSINESS
#    (take the latest census snapshot for each business_id)
biz <- df %>%
  group_by(business_id) %>%
  slice_max(census_year, n = 1, with_ties = FALSE) %>%
  ungroup()

# 3) (Optional, recommended) remove bias: keep only unbiased rows
#    i.e., closed businesses + long-lived (>12 yrs) per your rule
biz <- biz %>% filter(bias_filter == FALSE)

# 4) Keep the two industries and prep labels
biz2 <- biz %>%
  filter(industry %in% c("Cafes", "Takeaway")) %>%
  mutate(
    Industry = recode(industry,
                      "Cafes"    = "Restaurant",
                      "Takeaway" = "Takeaway")
  ) %>%
  # ensure clean target variable
  filter(!is.na(lifespan_years), lifespan_years >= 0)

# 5) Violin plot (with boxplot + median dot)
ggplot(biz2, aes(x = Industry, y = lifespan_years, fill = Industry)) +
  geom_violin(trim = TRUE, alpha = 0.85) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.6) +
  stat_summary(fun = median, geom = "point", size = 2.5, color = "black") +
  labs(
    title = "Distribution of Business Lifespan by Industry",
    subtitle = "Each business counted once (latest census record), bias-corrected sample",
    x = NULL, y = "Lifespan (years)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

```

```{r}


# 1️⃣ Load data
df <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)%>%
  filter(start_year_filter == TRUE)

# 2️⃣ Keep one record per business (latest census year, bias-filtered)
biz_unique <- df %>%
  group_by(business_id) %>%
  slice_max(census_year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(bias_filter == FALSE)

# 3️⃣ Select only Cafes and Takeaway
lifespan_by_area <- biz_unique %>%
  filter(industry %in% c("Cafes", "Takeaway")) %>%
  mutate(
    Industry = recode(industry,
                      "Cafes" = "Restaurant",
                      "Takeaway" = "Takeaway"),
    CLUE_Area = clue_area
  ) %>%
  filter(!is.na(lifespan_years) & !is.na(CLUE_Area)) %>%
  mutate(
    # ✅ Updated lifespan grouping
    lifespan_group = case_when(
      lifespan_years <= 3  ~ "0–3 yrs",
      lifespan_years <= 6  ~ "4–6 yrs",
      lifespan_years <= 9  ~ "7–9 yrs",
      lifespan_years <= 12 ~ "10–12 yrs",
      TRUE ~ ">12 yrs"
    ),
    lifespan_group = factor(lifespan_group,
                            levels = c("0–3 yrs", "4–6 yrs", "7–9 yrs", "10–12 yrs", ">12 yrs"))
  ) %>%
  group_by(CLUE_Area, Industry, lifespan_group) %>%
  summarise(Count = n(), .groups = "drop")

# 4️⃣ Plot with ordered x-axis
ggplot(lifespan_by_area, aes(x = lifespan_group, y = Count, fill = Industry)) +
  geom_col(position = "dodge") +
  facet_wrap(~ CLUE_Area, scales = "free_y") +
  labs(
    title = "Business Lifespan Distribution by CLUE Area",
    subtitle = "Restaurants vs Takeaway Food Services",
    x = "Lifespan Group",
    y = "Number of Businesses",
    fill = "Industry"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )

```

```{r}




# Load data
df_full <- read_csv("df_ULTRA_enhanced_with_traffic.csv")

# Bias-free cohort (closed OR veteran) AND valid start (first_seen > 2002 OR veteran)
df_analysis <- df_full %>%
  filter(bias_filter == FALSE, start_year_filter == TRUE) %>%   # ← add this condition
  arrange(business_id, census_year) %>%
  group_by(business_id) %>%
  slice_head(n = 1) %>%   # first observation per business
  ungroup()



# CLUE AREA CLUSTERING



# Aggregate metrics by area
area_metrics <- df_analysis %>%
  group_by(clue_area) %>%
  summarise(
    # Sample size
    n_businesses = n(),
    
    # Survival metrics
    mean_lifespan = mean(lifespan_years),
    median_lifespan = median(lifespan_years),
    success_rate = mean(lifespan_group == ">12 yrs") * 100,
    
    # Competition metrics
    mean_same_ind = mean(same_industry_200m, na.rm = TRUE),
    mean_other_ind = mean(other_industry_200m, na.rm = TRUE),
    mean_total_comp = mean(competitors_200m, na.rm = TRUE),
    
    # Business characteristics
    mean_seats = mean(num_seats, na.rm = TRUE),
    pct_cafe = mean(industry == "Cafes") * 100,
    pct_outdoor = mean(str_detect(seating_type, "Outdoor")) * 100,
    
    # Variability
    sd_lifespan = sd(lifespan_years),
    
    .groups = "drop"
  ) %>%
  filter(n_businesses >= 20)  # Only areas with sufficient data



# Prepare clustering features (standardized)
area_features <- area_metrics %>%
  dplyr::select(
    mean_lifespan,
    success_rate,
    mean_total_comp,
    mean_seats,
    sd_lifespan
  ) %>%
  scale() %>%
  as.data.frame()

rownames(area_features) <- area_metrics$clue_area


# Determine optimal number of clusters




# Elbow method
set.seed(42)
wss <- sapply(2:8, function(k) {
  kmeans(area_features, centers = k, nstart = 25)$tot.withinss
})

# Silhouette method
sil_widths <- sapply(2:8, function(k) {
  km <- kmeans(area_features, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist(area_features))
  mean(ss[, 3])
})

# Find optimal
optimal_k <- which.max(sil_widths) + 1
cat("Optimal clusters (silhouette method):", optimal_k, "\n\n")

# Create elbow + silhouette plots
p_elbow <- data.frame(k = 2:8, wss = wss) %>%
  ggplot(aes(x = k, y = wss)) +
  geom_line(color = "#3498db", size = 1.2) +
  geom_point(color = "#3498db", size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "red") +
  labs(title = "Elbow Method", x = "Number of Clusters", y = "Within-cluster SS") +
  theme_minimal(base_size = 11)

p_silhouette <- data.frame(k = 2:8, silhouette = sil_widths) %>%
  ggplot(aes(x = k, y = silhouette)) +
  geom_line(color = "#2ecc71", size = 1.2) +
  geom_point(color = "#2ecc71", size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "red") +
  labs(title = "Silhouette Analysis", x = "Number of Clusters", 
       y = "Average Silhouette Width") +
  theme_minimal(base_size = 11)

# Perform clustering


cat("Performing k-means clustering with k =", optimal_k, "\n")

set.seed(42)
km_areas <- kmeans(area_features, centers = optimal_k, nstart = 50)

# Add cluster assignments
area_metrics$cluster <- km_areas$cluster

# Order clusters by performance (best = 1, worst = highest number)
cluster_performance <- area_metrics %>%
  group_by(cluster) %>%
  summarise(avg_success = mean(success_rate), .groups = "drop") %>%
  arrange(desc(avg_success)) %>%
  mutate(rank = row_number())

area_metrics <- area_metrics %>%
  left_join(cluster_performance %>% dplyr::select(cluster, rank), by = "cluster") %>%
  mutate(cluster_ordered = rank) %>%
  dplyr::select(-rank)

# Assign tier names
tier_names <- c("Tier 1: Prime Locations", "Tier 2: Solid Locations", 
                "Tier 3: Moderate Locations", "Tier 4: Challenging Locations")
tier_names <- tier_names[1:optimal_k]

area_metrics <- area_metrics %>%
  mutate(tier = tier_names[cluster_ordered])

# Cluster Profiles


cluster_summary <- area_metrics %>%
  group_by(tier, cluster_ordered) %>%
  summarise(
    n_areas = n(),
    n_businesses = sum(n_businesses),
    avg_lifespan = round(mean(mean_lifespan), 1),
    avg_success_rate = round(mean(success_rate), 1),
    avg_competition = round(mean(mean_total_comp), 0),
    areas = paste(clue_area, collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(cluster_ordered)

print(cluster_summary)
cat("\n")





# Visualizations


# PCA for 2D visualization
pca_result <- prcomp(area_features)
pca_data <- as.data.frame(pca_result$x[, 1:2])
pca_data$cluster <- factor(area_metrics$cluster_ordered)
pca_data$tier <- area_metrics$tier
pca_data$area <- area_metrics$clue_area
pca_data$success_rate <- area_metrics$success_rate

var_explained <- summary(pca_result)$importance[2, 1:2] * 100

p_pca <- ggplot(pca_data, aes(x = PC1, y = PC2, color = tier, size = success_rate)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = area), size = 2.5, hjust = 0, vjust = 0, 
            nudge_x = 0.1, nudge_y = 0.1, check_overlap = TRUE) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(3, 10)) +
  labs(
    title = "CLUE Area Clustering (PCA Visualization)",
    subtitle = "Size = Success Rate | Color = Performance Tier",
    x = paste0("PC1 (", round(var_explained[1], 1), "% variance)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "% variance)"),
    color = "Tier",
    size = "Success Rate (%)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

# Hierarchical clustering dendrogram
dist_matrix <- dist(area_features)
hc <- hclust(dist_matrix, method = "ward.D2")

dend_data <- dendro_data(hc)
dend_data$labels$label <- area_metrics$clue_area[as.numeric(dend_data$labels$label)]
dend_data$labels$tier <- area_metrics$tier[as.numeric(dend_data$labels$label)]

p_dendrogram <- ggplot(segment(dend_data)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = label(dend_data), 
            aes(x = x, y = y, label = label, color = tier),
            hjust = 1, angle = 90, size = 3) +
  geom_hline(yintercept = hc$height[length(hc$height) - optimal_k + 2], 
             linetype = "dashed", color = "red", size = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Hierarchical Clustering: CLUE Areas",
    subtitle = paste("Dendrogram with", optimal_k, "clusters (red line)"),
    x = NULL,
    y = "Distance"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )

# Heatmap of cluster characteristics
cluster_means <- area_metrics %>%
  group_by(tier) %>%
  summarise(
    `Mean Lifespan` = mean(mean_lifespan),
    `Success Rate (%)` = mean(success_rate),
    `Competition` = mean(mean_total_comp),
    `Restaurant Size` = mean(mean_seats),
    `Lifespan Variability` = mean(sd_lifespan),
    .groups = "drop"
  ) %>%
  pivot_longer(-tier, names_to = "Metric", values_to = "Value") %>%
  group_by(Metric) %>%
  mutate(Value_scaled = scale(Value)[,1]) %>%
  ungroup()

p_heatmap <- ggplot(cluster_means, aes(x = tier, y = Metric, fill = Value_scaled)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = round(Value, 1)), size = 3, fontface = "bold") +
  scale_fill_gradient2(low = "#e74c3c", mid = "white", high = "#27ae60",
                       midpoint = 0, name = "Standardized\nValue") +
  labs(
    title = "Cluster Characteristics Heatmap",
    subtitle = "Numbers show actual values | Colors show relative performance",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    panel.grid = element_blank()
  )



#  STREET-LEVEL CLUSTERING 




# Aggregate metrics by street - CREATE UNIQUE ID
street_metrics <- df_analysis %>%
  group_by(street_address, clue_area) %>%
  summarise(
    # Sample size
    n_businesses = n(),
    
    # Historical performance
    mean_lifespan = mean(lifespan_years),
    median_lifespan = median(lifespan_years),
    success_rate = mean(lifespan_group == ">12 yrs") * 100,
    
    # Competition
    mean_same_ind = mean(same_industry_200m, na.rm = TRUE),
    mean_other_ind = mean(other_industry_200m, na.rm = TRUE),
    mean_competition = mean(competitors_200m, na.rm = TRUE),
    
    # Business characteristics
    mean_seats = mean(num_seats, na.rm = TRUE),
    pct_cafe = mean(industry == "Cafes") * 100,
    pct_outdoor = mean(str_detect(seating_type, "Outdoor"), na.rm = TRUE) * 100,
    
    # Variability
    sd_lifespan = sd(lifespan_years),
    
    .groups = "drop"
  ) %>%
  filter(n_businesses >= 4) %>%  # Only streets with sufficient history
  mutate(
    # CREATE UNIQUE ID: street_area combination
    street_id = paste0(street_address, " (", clue_area, ")")
  )



# Prepare clustering features (standardized)
street_features <- street_metrics %>%
  dplyr::select(
    mean_lifespan,
    success_rate,
    mean_competition
  ) %>%
  scale() %>%
  as.data.frame()

# Use unique street_id as row names
rownames(street_features) <- street_metrics$street_id


# Determine optimal clusters




set.seed(42)
street_sil <- sapply(2:6, function(k) {
  km <- kmeans(street_features, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist(street_features))
  mean(ss[, 3])
})

optimal_k_street <- which.max(street_sil) + 1
cat("Optimal clusters:", optimal_k_street, "\n\n")

# Perform clustering




set.seed(42)
km_streets <- kmeans(street_features, centers = optimal_k_street, nstart = 50)

street_metrics$cluster <- km_streets$cluster

# Order clusters by performance
street_cluster_perf <- street_metrics %>%
  group_by(cluster) %>%
  summarise(avg_success = mean(success_rate), .groups = "drop") %>%
  arrange(desc(avg_success)) %>%
  mutate(rank = row_number())

street_metrics <- street_metrics %>%
  left_join(street_cluster_perf %>% dplyr::select(cluster, rank), by = "cluster") %>%
  mutate(cluster_ordered = rank) %>%
  dplyr::select(-rank)

# Assign tier names
street_tier_names <- c("Gold Streets", "Silver Streets", "Bronze Streets", 
                       "Copper Streets", "Avoid Streets")
street_tier_names <- street_tier_names[1:optimal_k_street]

street_metrics <- street_metrics %>%
  mutate(street_tier = street_tier_names[cluster_ordered])


# Street Cluster Profiles




street_cluster_summary <- street_metrics %>%
  group_by(street_tier, cluster_ordered) %>%
  summarise(
    n_streets = n(),
    n_businesses = sum(n_businesses),
    avg_lifespan = round(mean(mean_lifespan), 1),
    avg_success_rate = round(mean(success_rate), 1),
    avg_competition = round(mean(mean_competition), 0),
    .groups = "drop"
  ) %>%
  arrange(cluster_ordered)

print(street_cluster_summary)
cat("\n")

# Top 10 streets per tier
for (i in 1:optimal_k_street) {
  tier_streets <- street_metrics %>%
    filter(cluster_ordered == i) %>%
    arrange(desc(success_rate)) %>%
    head(10)
  
  cat(street_tier_names[i], "\n")
  cat(strrep("─", 70), "\n")
  cat("Characteristics:\n")
  cat("  Avg lifespan:", round(mean(tier_streets$mean_lifespan), 1), "years\n")
  cat("  Avg success rate:", round(mean(tier_streets$success_rate), 1), "%\n")
  cat("  Avg competition:", round(mean(tier_streets$mean_competition), 0), "\n\n")
  cat("Top 10 streets:\n")
  for (j in 1:min(10, nrow(tier_streets))) {
    cat(sprintf("  %2d. %-50s - %.1f yrs, %.0f%% success\n",
                j, tier_streets$street_id[j],
                tier_streets$mean_lifespan[j], tier_streets$success_rate[j]))
  }
  cat("\n")
}


# Street Visualizations

# Scatter plot: Success Rate vs Competition
p_street_scatter <- ggplot(street_metrics, 
                           aes(x = mean_competition, y = success_rate, 
                               color = street_tier, size = n_businesses)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(
    values = c("Gold Streets" = "#FFD700", 
               "Silver Streets" = "#C0C0C0", 
               "Bronze Streets" = "#CD7F32", 
               "Copper Streets" = "#B87333", 
               "Avoid Streets" = "#8B0000")
  ) +
  scale_size_continuous(range = c(1, 8)) +
  labs(
    title = "Street Performance: Success Rate vs Competition",
    subtitle = "Size = Number of businesses | Color = Performance tier",
    x = "Average Competition (200m radius)",
    y = "Success Rate (%)",
    color = "Street Tier",
    size = "Businesses"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

# Distribution of streets across tiers
p_street_dist <- street_metrics %>%
  ggplot(aes(x = fct_reorder(street_tier, cluster_ordered), fill = street_tier)) +
  geom_bar(alpha = 0.8) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, fontface = "bold") +
  scale_fill_manual(
    values = c("Gold Streets" = "#FFD700", 
               "Silver Streets" = "#C0C0C0", 
               "Bronze Streets" = "#CD7F32", 
               "Copper Streets" = "#B87333", 
               "Avoid Streets" = "#8B0000")
  ) +
  labs(
    title = "Distribution of Streets Across Performance Tiers",
    x = NULL,
    y = "Number of Streets"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))

# Box plot: Lifespan distribution by tier
p_street_boxplot <- ggplot(street_metrics, 
                            aes(x = fct_reorder(street_tier, cluster_ordered), 
                                y = mean_lifespan, fill = street_tier)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
  scale_fill_manual(
    values = c("Gold Streets" = "#FFD700", 
               "Silver Streets" = "#C0C0C0", 
               "Bronze Streets" = "#CD7F32", 
               "Copper Streets" = "#B87333", 
               "Avoid Streets" = "#8B0000")
  ) +
  labs(
    title = "Lifespan Distribution by Street Tier",
    x = NULL,
    y = "Mean Lifespan (years)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))

# Heatmap: Top 20 streets by tier
top_streets <- street_metrics %>%
  group_by(street_tier) %>%
  arrange(desc(success_rate)) %>%
  slice_head(n = 4) %>%
  ungroup() %>%
  arrange(cluster_ordered, desc(success_rate))

p_street_heatmap <- top_streets %>%
  dplyr::select(street_id, street_tier, mean_lifespan, success_rate, mean_competition) %>%
  pivot_longer(cols = c(mean_lifespan, success_rate, mean_competition),
               names_to = "metric", values_to = "value") %>%
  mutate(
    metric = case_when(
      metric == "mean_lifespan" ~ "Avg Lifespan (yrs)",
      metric == "success_rate" ~ "Success Rate (%)",
      metric == "mean_competition" ~ "Competition"
    )
  ) %>%
  group_by(metric) %>%
  mutate(value_scaled = scale(value)[,1]) %>%
  ungroup() %>%
  ggplot(aes(x = metric, y = fct_reorder(street_id, as.numeric(factor(street_tier))), 
             fill = value_scaled)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = round(value, 1)), size = 2.5) +
  scale_fill_gradient2(low = "#e74c3c", mid = "white", high = "#27ae60",
                       midpoint = 0, name = "Standardized\nValue") +
  facet_grid(street_tier ~ ., scales = "free_y", space = "free_y") +
  labs(
    title = "Top Streets by Tier: Performance Metrics",
    subtitle = "Top 4 streets from each tier",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 20, hjust = 1),
    strip.text = element_text(face = "bold"),
    panel.grid = element_blank()
  )


# Save street plots


p_street_combined <- (p_street_scatter | p_street_boxplot) / 
                     p_street_dist / 
                     p_street_heatmap +
  plot_layout(heights = c(1.5, 0.8, 1.2)) +
  plot_annotation(
    title = "Street-Level Performance Clustering",
    subtitle = paste("Identified", optimal_k_street, "performance tiers across", 
                     nrow(street_metrics), "unique street-area combinations"),
    theme = theme(plot.title = element_text(face = "bold", size = 16))
  )

ggsave("clustering_streets_analysis.png", p_street_combined,
       width = 14, height = 14, dpi = 300, bg = "white")



# Save street metrics with tier assignments


write_csv(street_metrics, "clustering_streets.csv")


write_csv(street_cluster_summary, "clustering_street_summary.csv")



# SAVE ALL OUTPUTS


# Save cluster assignments
write_csv(area_metrics, "clustering_clue_areas.csv")

write_csv(street_metrics, "clustering_streets.csv")

write_csv(cluster_summary, "clustering_area_summary.csv")

write_csv(street_cluster_summary, "clustering_street_summary.csv")


# Save area plots
p_area_combined <- (p_elbow + p_silhouette) / p_pca / p_heatmap +
  plot_layout(heights = c(1, 2, 1.5)) +
  plot_annotation(
    title = "CLUE Area Clustering Analysis",
    subtitle = paste("Identified", optimal_k, "distinct performance tiers"),
    theme = theme(plot.title = element_text(face = "bold", size = 16))
  )

ggsave("clustering_areas_analysis.png", p_area_combined,
       width = 14, height = 16, dpi = 300, bg = "white")


ggsave("clustering_areas_dendrogram.png", p_dendrogram,
       width = 12, height = 8, dpi = 300, bg = "white")


# Save street plots
p_street_combined <- p_street_scatter / p_street_dist +
  plot_annotation(
    title = "Street-Level Performance Clustering",
    subtitle = paste("Identified", optimal_k_street, "performance tiers across", 
                     nrow(street_metrics), "streets"),
    theme = theme(plot.title = element_text(face = "bold", size = 16))
  )

ggsave("clustering_streets_analysis.png", p_street_combined,
       width = 12, height = 10, dpi = 300, bg = "white")



ggsave("pca.png", p_pca,
       width = 12, height = 10, dpi = 300, bg = "white")


ggsave("cluster.png", p_street_scatter,
       width = 12, height = 10, dpi = 300, bg = "white")

p_pca

p_dendrogram


p_heatmap

p_street_scatter



p_street_dist



```





```{r}


# 1) Area scatter with interpretable axes (mirrors street plot)
p_area_scatter <- ggplot(area_metrics,
  aes(x = mean_total_comp, y = success_rate,
      color = tier, size = n_businesses, label = clue_area)) +
  geom_point(alpha = 0.8) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 50, seed = 42, show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(3, 9), name = "Businesses") +
  labs(
    title = "CLUE Area Performance: Success Rate vs Competition",
    subtitle = "Color = cluster tier (k-means) | Size = # of businesses",
    x = "Average Competition (200m radius)",
    y = "Success Rate (%)",
    color = "Tier"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")

ggsave("clue_areas_competition_vs_success.png", p_area_scatter,
       width = 12, height = 9, dpi = 300, bg = "white")

# 2) (Optional) Alternative: Competition vs Mean Lifespan
p_area_scatter_life <- ggplot(area_metrics,
  aes(x = mean_total_comp, y = mean_lifespan,
      color = tier, size = n_businesses, label = clue_area)) +
  geom_point(alpha = 0.8) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 50, seed = 42, show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(3, 9), name = "Businesses") +
  labs(
    title = "CLUE Area Performance: Lifespan vs Competition",
    subtitle = "Color = cluster tier (k-means) | Size = # of businesses",
    x = "Average Competition (200m radius)",
    y = "Mean Lifespan (years)",
    color = "Tier"
  ) +
  theme_minimal(base_size = 11)

ggsave("clue_areas_competition_vs_lifespan.png", p_area_scatter_life,
       width = 12, height = 9, dpi = 300, bg = "white")



p_area_scatter_life

```



```{r}
library(tidyverse)
library(patchwork)
library(scales)
library(survival)
library(survminer)
library(tidyr)


# LOAD AND PREPARE DATA



df_full <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)

# Get completed lifecycles only (bias-free sample)
df_analysis <- df_full %>%
 filter(bias_filter == FALSE, start_year_filter == TRUE) %>%  
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  mutate(
    # Log transformation for size
    log_seats = log1p(num_seats),
    
    # Success indicator
    success = (lifespan_group == ">12 yrs"),
    
    # Business age categories
    age_category = case_when(
      lifespan_years <= 3 ~ "Early (1-3 yrs)",
      lifespan_years <= 8 ~ "Established (4-8 yrs)",
      lifespan_years <= 12 ~ "Mature (9-12 yrs)",
      TRUE ~ "Veteran (>12 yrs)"
    ),
    age_category = factor(age_category, levels = c("Early (1-3 yrs)", "Established (4-8 yrs)", 
                                                     "Mature (9-12 yrs)", "Veteran (>12 yrs)")),
    
    # Era categories
    era = case_when(
      census_year <= 2008 ~ "Pre-GFC (2002-2008)",
      census_year <= 2015 ~ "Recovery (2009-2015)",
      TRUE ~ "Modern (2016-2023)"
    ),
    
    # Competition levels
    comp_level = ifelse(other_industry_200m > median(other_industry_200m, na.rm=TRUE), 
                       "High Competition", "Low Competition"),
    
    # Size levels
    size_level = ifelse(num_seats > median(num_seats, na.rm=TRUE), 
                       "Large (>median)", "Small (≤median)")
  )


# Standardize features function
standardize_features <- function(df) {
  df %>%
    mutate(
      same_ind_std = scale(same_industry_200m)[,1],
      other_ind_std = scale(other_industry_200m)[,1],
      log_seats_std = scale(log_seats)[,1],
      foottraffic_std = scale(avg_daily_foottraffic)[,1]
    )
}

df_std <- standardize_features(df_analysis)

# Load street clustering and add street quality indicators to main dataset
street_clusters <- read_csv("clustering_streets.csv", show_col_types = FALSE)

df_std <- df_std %>%
  left_join(street_clusters %>% select(street_address, street_tier), 
            by = "street_address") %>%
  mutate(
    # Create best/worst street indicators
    on_best_street = as.numeric(street_tier %in% c("Gold Streets", "Silver Streets")),
    on_worst_street = as.numeric(street_tier %in% c("Copper Streets", "Avoid Streets"))
  )

# Create subset with valid street data for street-specific analyses
df_streets <- df_std %>%
  filter(!is.na(street_tier))


# ANALYSIS 1: GOOD VS POOR STREETS




# Define good vs poor streets (already have street_tier from earlier merge)
df_good_poor <- df_streets %>%
  mutate(
    street_quality = case_when(
      street_tier %in% c("Gold Streets") ~ "Prime Streets",
      street_tier %in% c("Copper Streets") ~ "Poor Streets",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(street_quality))



# Function to create comparison plots
create_street_comparison <- function(data, x_var, x_label, title_prefix) {
  
  # Calculate correlations by street quality
  cor_prime <- cor(data %>% filter(street_quality == "Prime Streets") %>% pull({{x_var}}),
                   data %>% filter(street_quality == "Prime Streets") %>% pull(lifespan_years),
                   use = "complete.obs")
  
  cor_poor <- cor(data %>% filter(street_quality == "Poor Streets") %>% pull({{x_var}}),
                  data %>% filter(street_quality == "Poor Streets") %>% pull(lifespan_years),
                  use = "complete.obs")
  
  ggplot(data, aes(x = {{x_var}}, y = lifespan_years, color = street_quality)) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(method = "lm", se = TRUE, size = 1.2) +
    scale_color_manual(values = c("Prime Streets" = "#27ae60", "Poor Streets" = "#e74c3c")) +
    annotate("text", x = -Inf, y = Inf, 
             label = paste0("Prime: r=", round(cor_prime, 3), "\nPoor: r=", round(cor_poor, 3)),
             hjust = -0.1, vjust = 1.2, size = 3, fontface = "bold") +
    labs(title = paste(title_prefix, "by Street Quality"),
         x = x_label, y = "Lifespan (years)",
         color = "Street Quality") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom",
          plot.title = element_text(face = "bold", size = 11))
}

p1_streets <- create_street_comparison(df_good_poor, same_ind_std, 
                                       "Same-Industry Competition", "A. Same-Industry")
p2_streets <- create_street_comparison(df_good_poor, other_ind_std, 
                                       "Other-Industry Competition", "B. Other-Industry")
p3_streets <- create_street_comparison(df_good_poor, log_seats_std, 
                                       "log(Seats)", "C. Restaurant Size")
p4_streets <- create_street_comparison(df_good_poor %>% filter(!is.na(foottraffic_std)), 
                                       foottraffic_std, "Pedestrians/day", "D. Foot Traffic")

p_streets_final <- (p1_streets | p2_streets) / (p3_streets | p4_streets) +
  plot_annotation(
    title = "Feature Effects: Prime vs Poor Streets",
    subtitle = "Do success factors work differently on good vs bad streets?",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 11, hjust = 0.5))
  )

ggsave("analysis1_good_vs_poor_streets.png", p_streets_final, 
       width = 14, height = 10, dpi = 300, bg = "white")



# Print insights
cat("KEY INSIGHTS:\n")
cor_diff_comp <- cor(df_good_poor %>% filter(street_quality == "Poor Streets") %>% 
                    pull(other_ind_std), 
                    df_good_poor %>% filter(street_quality == "Poor Streets") %>% 
                    pull(lifespan_years), use = "complete.obs") -
                cor(df_good_poor %>% filter(street_quality == "Prime Streets") %>% 
                    pull(other_ind_std), 
                    df_good_poor %>% filter(street_quality == "Prime Streets") %>% 
                    pull(lifespan_years), use = "complete.obs")

cat("  Competition hurts", ifelse(cor_diff_comp < 0, "MORE", "LESS"), 
    "on poor streets (Δr =", round(cor_diff_comp, 3), ")\n\n")


# ANALYSIS 2: BY CLUE AREA


# Load CLUE area clustering
clue_clusters <- read_csv("clustering_clue_areas.csv", show_col_types = FALSE)

# Select top 4 areas by sample size
top_areas <- df_std %>%
  count(clue_area, sort = TRUE) %>%
  head(4) %>%
  pull(clue_area)

df_areas <- df_std %>%
  filter(clue_area %in% top_areas)

cat("Top 4 CLUE areas:\n")
print(df_areas %>% count(clue_area, sort = TRUE))
cat("\n")

# Create small multiples for each feature
create_area_plot <- function(data, x_var, x_label, title) {
  
  # Calculate correlations by area
  cors <- data %>%
    group_by(clue_area) %>%
    summarise(
      r = cor({{x_var}}, lifespan_years, use = "complete.obs"),
      .groups = "drop"
    )
  
  ggplot(data, aes(x = {{x_var}}, y = lifespan_years, color = clue_area)) +
    geom_point(alpha = 0.2, size = 0.5) +
    geom_smooth(method = "lm", se = FALSE, size = 1) +
    facet_wrap(~clue_area, ncol = 2) +
    scale_color_brewer(palette = "Set1") +
    labs(title = title, x = x_label, y = "Lifespan (years)") +
    theme_minimal(base_size = 9) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", size = 10),
          strip.text = element_text(face = "bold"))
}

p1_areas <- create_area_plot(df_areas, same_ind_std, 
                             "Same-Industry Competition", "A. Same-Industry by Area")
p2_areas <- create_area_plot(df_areas, other_ind_std, 
                             "Other-Industry Competition", "B. Other-Industry by Area")
p3_areas <- create_area_plot(df_areas, log_seats_std, 
                             "log(Seats)", "C. Size by Area")
p4_areas <- create_area_plot(df_areas %>% filter(!is.na(foottraffic_std)), 
                             foottraffic_std, "Pedestrians/day", "D. Foot Traffic by Area")

p_areas_final <- (p1_areas | p2_areas) / (p3_areas | p4_areas) +
  plot_annotation(
    title = "Feature Effects by CLUE Area",
    subtitle = "Different neighborhoods, different rules?",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 11, hjust = 0.5))
  )

ggsave("analysis2_by_clue_area.png", p_areas_final, 
       width = 14, height = 12, dpi = 300, bg = "white")


# ANALYSIS 3: BUSINESS LIFECYCLE STAGE



# Calculate average lifespan and correlations by stage
lifecycle_summary <- df_std %>%
  group_by(age_category) %>%
  summarise(
    n = n(),
    avg_lifespan = mean(lifespan_years),
    cor_comp = cor(other_ind_std, lifespan_years, use = "complete.obs"),
    cor_size = cor(log_seats_std, lifespan_years, use = "complete.obs"),
    .groups = "drop"
  )

# Plot: Average lifespan by stage (sanity check)
p1_lifecycle <- ggplot(lifecycle_summary, aes(x = age_category, y = avg_lifespan, fill = age_category)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = round(avg_lifespan, 1)), vjust = -0.5, fontface = "bold") +
  scale_fill_brewer(palette = "RdYlGn") +
  labs(title = "A. Average Lifespan by Stage",
       x = NULL, y = "Average Lifespan (years)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))

# Plot: Competition effect by stage
p2_lifecycle <- ggplot(lifecycle_summary, aes(x = age_category, y = cor_comp, fill = age_category)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = round(cor_comp, 3)), vjust = -0.5, fontface = "bold") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_brewer(palette = "RdYlGn") +
  labs(title = "B. Competition Effect by Stage",
       x = NULL, y = "Correlation with Lifespan") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))

# Plot: Size effect by stage
p3_lifecycle <- ggplot(lifecycle_summary, aes(x = age_category, y = cor_size, fill = age_category)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = round(cor_size, 3)), vjust = -0.5, fontface = "bold") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_brewer(palette = "RdYlGn") +
  labs(title = "C. Size Effect by Stage",
       x = NULL, y = "Correlation with Lifespan") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))

# Scatter: Competition by stage
p4_lifecycle <- ggplot(df_std, aes(x = other_ind_std, y = lifespan_years, color = age_category)) +
  geom_point(alpha = 0.3, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_brewer(palette = "RdYlGn") +
  labs(title = "D. Competition Effects Across Stages",
       x = "Other-Industry Competition", y = "Lifespan (years)",
       color = "Stage") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))

p_lifecycle_final <- (p1_lifecycle | p2_lifecycle) / (p3_lifecycle | p4_lifecycle) +
  plot_annotation(
    title = "Business Lifecycle Stage Analysis",
    subtitle = "What matters in Year 1 vs Year 10?",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 11, hjust = 0.5))
  )

ggsave("analysis3_lifecycle_stages.png", p_lifecycle_final, 
       width = 14, height = 10, dpi = 300, bg = "white")




# ANALYSIS 4: INTERACTION EFFECTS


# Interaction 1: Competition × Size
interaction1 <- df_std %>%
  filter(!is.na(comp_level), !is.na(size_level)) %>%
  group_by(comp_level, size_level) %>%
  summarise(
    n = n(),
    avg_lifespan = mean(lifespan_years),
    success_rate = mean(success) * 100,
    .groups = "drop"
  )

p1_interact <- ggplot(interaction1, aes(x = comp_level, y = size_level, fill = avg_lifespan)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = sprintf("%.1f yrs\n(%d)", avg_lifespan, n)), 
            fontface = "bold", size = 4) +
  scale_fill_gradient2(low = "#e74c3c", mid = "#f39c12", high = "#27ae60",
                       midpoint = mean(df_std$lifespan_years)) +
  labs(title = "A. Competition × Size Interaction",
       x = "Competition Level", y = "Restaurant Size",
       fill = "Avg Lifespan") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# Interaction 2: Competition × Street Quality
interaction2 <- df_good_poor %>%
  group_by(comp_level, street_quality) %>%
  summarise(
    n = n(),
    avg_lifespan = mean(lifespan_years),
    .groups = "drop"
  )

p2_interact <- ggplot(interaction2, aes(x = comp_level, y = street_quality, fill = avg_lifespan)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = sprintf("%.1f yrs\n(%d)", avg_lifespan, n)), 
            fontface = "bold", size = 4) +
  scale_fill_gradient2(low = "#e74c3c", mid = "#f39c12", high = "#27ae60",
                       midpoint = mean(df_good_poor$lifespan_years)) +
  labs(title = "B. Competition × Street Quality",
       x = "Competition Level", y = "Street Quality",
       fill = "Avg Lifespan") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# Interaction 3: Size × Street Quality
interaction3 <- df_good_poor %>%
  group_by(size_level, street_quality) %>%
  summarise(
    n = n(),
    avg_lifespan = mean(lifespan_years),
    .groups = "drop"
  )

p3_interact <- ggplot(interaction3, aes(x = size_level, y = street_quality, fill = avg_lifespan)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = sprintf("%.1f yrs\n(%d)", avg_lifespan, n)), 
            fontface = "bold", size = 4) +
  scale_fill_gradient2(low = "#e74c3c", mid = "#f39c12", high = "#27ae60",
                       midpoint = mean(df_good_poor$lifespan_years)) +
  labs(title = "C. Size × Street Quality",
       x = "Restaurant Size", y = "Street Quality",
       fill = "Avg Lifespan") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# Line plot showing interaction visually
p4_interact <- ggplot(interaction1, aes(x = comp_level, y = avg_lifespan, 
                                        color = size_level, group = size_level)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  geom_text(aes(label = round(avg_lifespan, 1)), 
            vjust = -1, fontface = "bold", size = 3.5) +
  scale_color_manual(values = c("Large (>median)" = "#27ae60", 
                                 "Small (≤median)" = "#e74c3c")) +
  labs(title = "D. Does Size Buffer Competition?",
       x = "Competition Level", y = "Average Lifespan (years)",
       color = "Size") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))

p_interact_final <- (p1_interact | p2_interact) / (p3_interact | p4_interact) +
  plot_annotation(
    title = "Interaction Effects: Do Factors Work Together?",
    subtitle = "Testing if size/location can buffer competition effects",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 11, hjust = 0.5))
  )

ggsave("analysis4_interactions.png", p_interact_final, 
       width = 14, height = 10, dpi = 300, bg = "white")




# ANALYSIS 5: OPTIMAL RANGE ANALYSIS




# Function to create decile analysis
create_decile_plot <- function(data, var, var_label, title) {
  data %>%
    filter(!is.na({{var}})) %>%
    mutate(decile = ntile({{var}}, 10)) %>%
    group_by(decile) %>%
    summarise(
      n = n(),
      avg_value = mean({{var}}, na.rm = TRUE),
      avg_lifespan = mean(lifespan_years, na.rm = TRUE),
      success_rate = mean(success, na.rm = TRUE) * 100,
      .groups = "drop"
    ) %>%
    ggplot(aes(x = decile, y = avg_lifespan)) +
    geom_line(color = "#3498db", size = 1.5) +
    geom_point(size = 3, color = "#3498db") +
    geom_hline(yintercept = mean(data$lifespan_years, na.rm = TRUE), 
               linetype = "dashed", color = "red") +
    scale_x_continuous(breaks = 1:10) +
    labs(title = title,
         x = paste(var_label, "Decile (1=lowest, 10=highest)"),
         y = "Average Lifespan (years)") +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(face = "bold"))
}

p1_optimal <- create_decile_plot(df_std, num_seats, "Restaurant Size", 
                                 "A. Size Sweet Spot (Inverted-U?)")
p2_optimal <- create_decile_plot(df_std, other_industry_200m, "Other-Industry Competition", 
                                 "B. Competition Effect (Linear Decline?)")
p3_optimal <- create_decile_plot(df_std, same_industry_200m, "Same-Industry Competition", 
                                 "C. Same-Industry Effect")
p4_optimal <- create_decile_plot(df_std %>% filter(!is.na(avg_daily_foottraffic)), 
                                 avg_daily_foottraffic, "Foot Traffic", 
                                 "D. Foot Traffic Effect")

p_optimal_final <- (p1_optimal | p2_optimal) / (p3_optimal | p4_optimal) +
  plot_annotation(
    title = "Optimal Range Analysis: Finding Sweet Spots",
    subtitle = "Red line = overall average | Looking for non-linear patterns",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 11, hjust = 0.5))
  )

ggsave("analysis5_optimal_ranges.png", p_optimal_final, 
       width = 14, height = 10, dpi = 300, bg = "white")




# ANALYSIS 6: SUCCESS PROFILE (12+ YEAR SURVIVORS)





# Compare successful vs failed businesses
success_profile <- df_std %>%
  mutate(outcome = case_when(
    lifespan_years <= 3 ~ "Failed (<3 yrs)",
    lifespan_years > 12 ~ "Success (>12 yrs)",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(outcome)) %>%
  group_by(outcome) %>%
  summarise(
    n = n(),
    avg_same_comp = mean(same_industry_200m, na.rm = TRUE),
    avg_other_comp = mean(other_industry_200m, na.rm = TRUE),
    avg_seats = mean(num_seats, na.rm = TRUE),
    pct_cbd = mean(clue_area == "Melbourne (CBD)", na.rm = TRUE) * 100,
    pct_cafe = mean(industry == "Cafes", na.rm = TRUE) * 100,
    .groups = "drop"
  )

# Replace NA values with 0 for percentage calculations
success_profile <- success_profile %>%
  mutate(across(starts_with("pct_"), ~replace_na(., 0)))


print(success_profile)


# Reshape for plotting
success_long <- success_profile %>%
  select(-n) %>%
  pivot_longer(-outcome, names_to = "metric", values_to = "value") %>%
  mutate(
    metric_clean = case_when(
      metric == "avg_same_comp" ~ "Same-Industry\nCompetition",
      metric == "avg_other_comp" ~ "Other-Industry\nCompetition",
      metric == "avg_seats" ~ "Restaurant\nSize (seats)",
      metric == "pct_cbd" ~ "% in CBD",
      metric == "pct_cafe" ~ "% Cafes",
      TRUE ~ metric
    )
  )

# Plot: Success vs Failed comparison
p_success <- ggplot(success_long, aes(x = metric_clean, y = value, fill = outcome)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(value, 1)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 3, fontface = "bold") +
  scale_fill_manual(values = c("Failed (<3 yrs)" = "#e74c3c", 
                                "Success (>12 yrs)" = "#27ae60")) +
  labs(title = "Success Profile: What Do Long-Term Survivors Have in Common?",
       subtitle = paste0("Comparing ", success_profile$n[1], " failed businesses vs ", 
                        success_profile$n[2], " successful businesses"),
       x = NULL, y = "Value",
       fill = "Outcome") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        plot.subtitle = element_text(size = 10, hjust = 0.5),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9))

ggsave("analysis6_success_profile.png", p_success, 
       width = 14, height = 8, dpi = 300, bg = "white")



# Calculate odds ratios
odds_prime_success <- (success_profile$pct_prime_street[2] / (100 - success_profile$pct_prime_street[2])) /
                      (success_profile$pct_prime_street[1] / (100 - success_profile$pct_prime_street[1]))




# ANALYSIS 7: SURVIVAL CURVES BY STREET TIER






# Prepare survival data
df_survival <- df_streets %>%
  filter(!is.na(street_tier)) %>%
  mutate(
    # Event indicator (1 = observed death, 0 = censored)
    event = 1,  # All are completed lifecycles, so all died
    time = lifespan_years
  )

cat("Sample sizes by street tier:\n")
print(df_survival %>% count(street_tier))
cat("\n")

# Create survival object
surv_obj <- Surv(time = df_survival$time, event = df_survival$event)

# Fit survival curves by street tier
fit_streets <- survfit(surv_obj ~ street_tier, data = df_survival)

# Plot survival curves
p_survival <- ggsurvplot(
  fit_streets,
  data = df_survival,
  conf.int = TRUE,
  risk.table = FALSE,
  pval = TRUE,
  pval.coord = c(0, 0.1),
  palette = c("Gold Streets" = "#FFD700",
              "Silver Streets" = "#C0C0C0",
              "Bronze Streets" = "#CD7F32",
              "Copper Streets" = "#B87333",
              "Avoid Streets" = "#8B0000"),
  legend.title = "Street Tier",
  legend.labs = levels(factor(df_survival$street_tier)),
  xlab = "Years in Business",
  ylab = "Survival Probability",
  title = "Survival Curves by Street Tier",
  subtitle = "How long do restaurants survive on different street tiers?",
  ggtheme = theme_minimal(base_size = 12),
  font.main = c(16, "bold"),
  font.subtitle = c(11, "plain"),
  font.x = c(12, "plain"),
  font.y = c(12, "plain")
)

# Save survival plot
ggsave("analysis7_survival_curves.png", print(p_survival), 
       width = 12, height = 8, dpi = 300, bg = "white")


# Calculate median survival times
median_survival <- summary(fit_streets)$table[, "median"]

print(median_survival)







# Display all plots
print(p_streets_final)
print(p_areas_final)
print(p_lifecycle_final)
print(p_interact_final)
print(p_optimal_final)
print(p_success)
print(p_survival)
```














#### traditional RM
```{r}

library(tidyverse)
library(randomForest)
library(caret)
library(ranger)      
library(patchwork)
library(scales)




df_shiny <- read_csv("df_clean_200m_shinny.csv", show_col_types = FALSE)
df_full  <- read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)

#  Select only columns needed for merging
filter_info <- df_full %>%
  select(business_id, start_year_filter) %>%
  distinct()

#  Join the new filter into the Shiny dataset
df_shiny_updated <- df_shiny %>%
  left_join(filter_info, by = "business_id")%>%
  mutate(start_year_filter = replace_na(start_year_filter, FALSE))

#  Sanity check — make sure row count remains the same
stopifnot(nrow(df_shiny_updated) == nrow(df_shiny))

#  Save updated dataset
write_csv(df_shiny_updated, "df_clean_200m_shinny.csv")


# STEP 1: LOAD AND PREPARE DATA


# Load modeling data
df_model <- read_csv("df_clean_200m_shinny.csv", show_col_types = FALSE)%>%

dplyr::filter(start_year_filter == TRUE)

cat("Dataset loaded:\n")
cat("  Rows:", nrow(df_model), "\n")
cat("  Features:", ncol(df_model) - 2, "(excluding business_id and target)\n\n")

# Remove business_id (not a predictor)
df_features <- df_model %>%
  dplyr::select(-business_id,-on_best_street,-on_worst_street, start_year_filter)

# Check for missing values
na_summary <- colSums(is.na(df_features))
if (sum(na_summary) > 0) {
  cat("⚠ WARNING: Missing values detected:\n")
  print(na_summary[na_summary > 0])
  cat("\nRemoving rows with missing values...\n")
  df_features <- na.omit(df_features)
}

cat("Final dataset:\n")
cat("  Complete cases:", nrow(df_features), "\n")
cat("  Target variable: lifespan_numeric\n")
cat("  Range:", min(df_features$lifespan_numeric), "-", 
    max(df_features$lifespan_numeric), "years\n\n")


# STEP 2: TRAIN/TEST SPLIT (80/20)




set.seed(42)
train_index <- createDataPartition(
  df_features$lifespan_numeric, 
  p = 0.80,           # 80% training
  list = FALSE,
  times = 1
)

df_train <- df_features[train_index, ]
df_test <- df_features[-train_index, ]



# STEP 3: FEATURE STANDARDIZATION (FOR COMPARISON)


# Identify numeric features (exclude target)
numeric_features <- df_train %>%
  dplyr::select(-lifespan_numeric) %>%
  dplyr::select(where(is.numeric)) %>%
  names()



# Calculate standardization parameters from TRAINING set only
preproc_params <- preProcess(
  df_train[, numeric_features], 
  method = c("center", "scale")
)

# Apply to both train and test
df_train_scaled <- df_train
df_train_scaled[, numeric_features] <- predict(
  preproc_params, 
  df_train[, numeric_features]
)

df_test_scaled <- df_test
df_test_scaled[, numeric_features] <- predict(
  preproc_params, 
  df_test[, numeric_features]
)


# STEP 4: CROSS-VALIDATION SETUP


# 5-fold CV with consistent folds
set.seed(42)
cv_folds <- createFolds(
  df_train$lifespan_numeric, 
  k = 5, 
  returnTrain = TRUE
)

cv_control <- trainControl(
  method = "cv",
  number = 5,
  verboseIter = TRUE,
  savePredictions = "final",
  index = cv_folds,
  allowParallel = TRUE  # Enable parallel processing if available
)


# STEP 5: HYPERPARAMETER TUNING GRID


# Define tuning grid
# For ranger (faster implementation)
n_features <- ncol(df_train) - 1  # Exclude target

tune_grid <- expand.grid(
  mtry = c(3, 5, 7, 9, 11),                          # Features per split
  splitrule = c("variance", "extratrees"),           # Splitting rule
  min.node.size = c(5, 10, 20, 30)                   # Min samples per leaf
)


# STEP 6: TRAIN MODELS (NON-STANDARDIZED VS STANDARDIZED)

# MODEL 1: Non-Standardized (Standard for Random Forest)




set.seed(42)
start_time_1 <- Sys.time()

rf_model_raw <- train(
  lifespan_numeric ~ .,
  data = df_train,
  method = "ranger",
  trControl = cv_control,
  tuneGrid = tune_grid,
  num.trees = 500,
  importance = "impurity",
  num.threads = 4  # Adjust based on your CPU
)

elapsed_1 <- round(difftime(Sys.time(), start_time_1, units = "mins"), 1)

cat("\n✓ Model 1 training complete (", elapsed_1, "minutes)\n\n")
cat("Best hyperparameters (non-standardized):\n")
print(rf_model_raw$bestTune)
cat("\nCV Performance:\n")
print(rf_model_raw$results %>% 
        filter(
          mtry == rf_model_raw$bestTune$mtry,
          splitrule == rf_model_raw$bestTune$splitrule,
          min.node.size == rf_model_raw$bestTune$min.node.size
        ) %>%
        dplyr::select(RMSE, Rsquared, MAE))


# MODEL 2: Standardized Features (for comparison)



set.seed(42)
start_time_2 <- Sys.time()

rf_model_scaled <- train(
  lifespan_numeric ~ .,
  data = df_train_scaled,
  method = "ranger",
  trControl = cv_control,
  tuneGrid = tune_grid,
  num.trees = 500,
  importance = "impurity",
  num.threads = 4
)

elapsed_2 <- round(difftime(Sys.time(), start_time_2, units = "mins"), 1)

cat("\n✓ Model 2 training complete (", elapsed_2, "minutes)\n\n")
cat("Best hyperparameters (standardized):\n")
print(rf_model_scaled$bestTune)
cat("\nCV Performance:\n")
print(rf_model_scaled$results %>% 
        filter(
          mtry == rf_model_scaled$bestTune$mtry,
          splitrule == rf_model_scaled$bestTune$splitrule,
          min.node.size == rf_model_scaled$bestTune$min.node.size
        ) %>%
        dplyr::select(RMSE, Rsquared, MAE))


# STEP 7: TEST SET EVALUATION



# Predictions on test set
pred_raw <- predict(rf_model_raw, df_test)
pred_scaled <- predict(rf_model_scaled, df_test_scaled)

# Calculate test metrics
test_metrics <- tibble(
  Model = c("Non-Standardized (RF Standard)", "Standardized (Comparison)"),
  
  `Test RMSE` = c(
    RMSE(pred_raw, df_test$lifespan_numeric),
    RMSE(pred_scaled, df_test$lifespan_numeric)
  ),
  
  `Test MAE` = c(
    MAE(pred_raw, df_test$lifespan_numeric),
    MAE(pred_scaled, df_test$lifespan_numeric)
  ),
  
  `Test R²` = c(
    R2(pred_raw, df_test$lifespan_numeric),
    R2(pred_scaled, df_test$lifespan_numeric)
  ),
  
  `CV RMSE` = c(
    min(rf_model_raw$results$RMSE),
    min(rf_model_scaled$results$RMSE)
  ),
  
  `CV R²` = c(
    max(rf_model_raw$results$Rsquared),
    max(rf_model_scaled$results$Rsquared)
  )
) %>%
  mutate(
    `Overfit Gap (CV-Test R²)` = `CV R²` - `Test R²`,
    `RMSE Diff (CV-Test)` = `CV RMSE` - `Test RMSE`
  )


print(test_metrics, width = Inf)

# Select best model
best_model <- if (test_metrics$`Test R²`[1] >= test_metrics$`Test R²`[2]) {
  cat("\n✓ WINNER: Non-Standardized model (as expected for Random Forest)\n")
  rf_model_raw
} else {
  cat("\n✓ WINNER: Standardized model (unexpected!)\n")
  rf_model_scaled
}


# STEP 8: FEATURE IMPORTANCE ANALYSIS




# Extract variable importance
importance_df <- varImp(best_model)$importance %>%
  rownames_to_column("Feature") %>%
  arrange(desc(Overall)) %>%
  mutate(
    Feature_Clean = case_when(
      Feature == "same_ind" ~ "Same-Industry Competition (200m)",
      Feature == "other_ind" ~ "Other-Industry Competition (200m)",
      Feature == "log_seats" ~ "Restaurant Size (log seats)",
      Feature == "on_best_street" ~ "Prime Location Street",
      Feature == "on_worst_street" ~ "Poor Quality Street",
      Feature == "property_competition" ~ "Property-Level Competition",
      Feature == "industry_cafe" ~ "Cafe Business Model",
      Feature == "outdoor_seating" ~ "Outdoor Seating",
      Feature == "mixed_seating" ~ "Mixed Seating Type",
      TRUE ~ str_to_title(str_replace_all(Feature, "_", " "))
    ),
    Rank = row_number()
  )


print(importance_df %>% 
        head(15) %>% 
        dplyr::select(Rank, Feature = Feature_Clean, Importance = Overall),
      n = 15)


# STEP 9: DIAGNOSTIC VISUALIZATIONS



# Choose best predictions for plots
best_pred <- if (test_metrics$`Test R²`[1] >= test_metrics$`Test R²`[2]) {
  pred_raw
} else {
  pred_scaled
}

# Plot 1: Predicted vs Actual
p1 <- data.frame(
  actual = df_test$lifespan_numeric,
  predicted = best_pred
) %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "#3498db") +
  geom_abline(slope = 1, intercept = 0, color = "#e74c3c", 
              linetype = "dashed", size = 1) +
  geom_smooth(method = "lm", color = "#2ecc71", fill = "#2ecc71", 
              alpha = 0.2, se = TRUE) +
  annotate("text", x = min(df_test$lifespan_numeric), 
           y = max(best_pred),
           label = paste0(
             "Test R² = ", round(test_metrics$`Test R²`[1], 3), "\n",
             "Test RMSE = ", round(test_metrics$`Test RMSE`[1], 2), " years\n",
             "Test MAE = ", round(test_metrics$`Test MAE`[1], 2), " years"
           ),
           hjust = 0, vjust = 1, size = 4, fontface = "bold") +
  labs(
    title = "Model Performance: Predicted vs Actual Lifespan",
    subtitle = "Test set evaluation (20% hold-out)",
    x = "Actual Lifespan (years)",
    y = "Predicted Lifespan (years)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

# Plot 2: Residuals
residuals <- df_test$lifespan_numeric - best_pred

p2 <- data.frame(
  predicted = best_pred,
  residuals = residuals
) %>%
  ggplot(aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.4, color = "#9b59b6") +
  geom_hline(yintercept = 0, color = "#e74c3c", linetype = "dashed", size = 1) +
  geom_smooth(method = "loess", color = "#2ecc71", fill = "#2ecc71", 
              alpha = 0.2, se = TRUE) +
  labs(
    title = "Residual Plot",
    subtitle = "Check for systematic bias",
    x = "Predicted Lifespan (years)",
    y = "Residuals (Actual - Predicted)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

# Plot 3: Feature Importance (Top 15)
p3 <- importance_df %>%
  head(15) %>%
  ggplot(aes(x = reorder(Feature_Clean, Overall), y = Overall)) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Feature Importance (Top 15)",
    subtitle = "Based on reduction in node impurity",
    x = NULL,
    y = "Importance Score"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# Plot 4: Hyperparameter Tuning Results
p4 <- rf_model_raw$results %>%
  ggplot(aes(x = mtry, y = Rsquared, color = factor(min.node.size))) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~splitrule) +
  scale_color_brewer(palette = "Set1", name = "Min Node Size") +
  labs(
    title = "Hyperparameter Tuning: R² vs mtry",
    subtitle = "Across different split rules and node sizes",
    x = "mtry (features per split)",
    y = "Cross-Validation R²"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")

# Combine plots
p_final <- (p1 | p2) / (p3 | p4) +
  plot_annotation(
    title = "Random Forest Model Diagnostics - Proper ML Workflow",
    subtitle = paste0(
      "5-Fold CV | Train/Test Split (80/20) | Hyperparameter Tuning | ",
      nrow(df_train), " training samples | ", nrow(df_test), " test samples"
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(size = 12)
    )
  )

# Save
ggsave("rf_model_proper_workflow.png", p_final, 
       width = 16, height = 12, dpi = 300, bg = "white")


# STEP 10: FINAL MODEL SUMMARY & EXPORT




best_test_metrics <- test_metrics %>% 
  filter(`Test R²` == max(`Test R²`)) %>%
  slice(1)



# Baseline comparison
baseline_rmse <- sqrt(mean((df_test$lifespan_numeric - mean(df_train$lifespan_numeric))^2))
improvement <- (baseline_rmse - best_test_metrics$`Test RMSE`) / baseline_rmse * 100



# Top 5 features

top5 <- importance_df %>% head(5)
for (i in 1:5) {
  cat(sprintf("  %d. %-45s (Importance: %.1f)\n", 
              i, top5$Feature_Clean[i], top5$Overall[i]))
}


# Save final model
saveRDS(best_model, "rf_model_final_tuned.rds")
saveRDS(preproc_params, "preprocessing_params.rds")



# Export predictions
predictions_export <- tibble(
  actual_lifespan = df_test$lifespan_numeric,
  predicted_lifespan = best_pred,
  residual = residuals,
  abs_error = abs(residuals)
) %>%
  arrange(desc(abs_error))

write_csv(predictions_export, "rf_test_predictions.csv")


# Export feature importance
write_csv(importance_df, "rf_feature_importance.csv")


# Display final plot

p1
p2
p3
p4

p_final
```






```{r}
library(tidyverse)
library(patchwork)
library(scales)


# LOAD DATA


df_full <- read_csv("df_ULTRA_enhanced_with_traffic.csv")

df_analysis <- df_full %>%
  filter(bias_filter == FALSE, start_year_filter == TRUE) %>%  
  group_by(business_id) %>%
  slice_head(n = 1) %>%
  ungroup()


# STANDARDIZE FEATURES (z-score: mean=0, sd=1)


df_standardized <- df_analysis %>%
  mutate(
    # Standardize competition metrics
    same_industry_std = scale(same_industry_200m)[,1],
    other_industry_std = scale(other_industry_200m)[,1],
    
    # Standardize size (use log transformation THEN standardize)
    log_seats = log1p(num_seats),
    log_seats_std = scale(log_seats)[,1],
    
    # Standardize foot traffic
    foottraffic_std = scale(avg_daily_foottraffic)[,1]
  )

# Check standardization




# Filter for foot traffic analysis
df_traffic_std <- df_standardized %>% filter(!is.na(foottraffic_std))


# CALCULATE CORRELATIONS (same as before - correlation is scale-invariant)


# Add use = "complete.obs" to handle NA values
cor_same <- cor(df_standardized$same_industry_std, df_standardized$lifespan_years, 
                use = "complete.obs")
cor_other <- cor(df_standardized$other_industry_std, df_standardized$lifespan_years, 
                 use = "complete.obs")
cor_size <- cor(df_standardized$log_seats_std, df_standardized$lifespan_years, 
                use = "complete.obs")
cor_traffic <- cor(df_traffic_std$foottraffic_std, df_traffic_std$lifespan_years, 
                   use = "complete.obs")

# Linear models (use na.action to handle missing values)
lm_same <- lm(lifespan_years ~ same_industry_std, 
              data = df_standardized, na.action = na.exclude)
lm_other <- lm(lifespan_years ~ other_industry_std, 
               data = df_standardized, na.action = na.exclude)
lm_size <- lm(lifespan_years ~ log_seats_std, 
              data = df_standardized, na.action = na.exclude)
lm_traffic <- lm(lifespan_years ~ foottraffic_std, 
                 data = df_traffic_std, na.action = na.exclude)




# CREATE STANDARDIZED PLOTS


# Panel A: Same-Industry Competition (Standardized)
p_same <- ggplot(df_standardized, aes(x = same_industry_std, y = lifespan_years)) +
  geom_point(alpha = 0.15, size = 1, color = "#3498db") +
  geom_smooth(method = "lm", color = "#e74c3c", fill = "#e74c3c", 
              alpha = 0.2, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", size = 0.5) +
  labs(title = "A. Same-Industry Competition",
       subtitle = "Clear negative relationship → Keep in model",
       x = "Competitors (200m)", y = "Lifespan (years)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", color = "#2c3e50"),
        plot.subtitle = element_text(color = "#27ae60"),
        panel.border = element_rect(color = "gray80", fill = NA, size = 0.8))

# Panel B: Other-Industry Competition (Standardized)
p_other <- ggplot(df_standardized, aes(x = other_industry_std, y = lifespan_years)) +
  geom_point(alpha = 0.15, size = 1, color = "#1abc9c") +
  geom_smooth(method = "lm", color = "#e74c3c", fill = "#e74c3c", 
              alpha = 0.2, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", size = 0.5) +
  labs(title = "B. Other-Industry Competition",
       subtitle = "Strong negative relationship → Keep in model",
       x = "Competitors (200m)", y = "Lifespan (years)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", color = "#2c3e50"),
        plot.subtitle = element_text(color = "#27ae60"),
        panel.border = element_rect(color = "gray80", fill = NA, size = 0.8))

# Panel C: Restaurant Size (Log-transformed THEN Standardized)
p_size <- ggplot(df_standardized, aes(x = log_seats_std, y = lifespan_years)) +
  geom_point(alpha = 0.15, size = 1, color = "#f39c12") +
  geom_smooth(method = "lm", color = "#e74c3c", fill = "#e74c3c", 
              alpha = 0.2, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", size = 0.5) +
  labs(title = "C. Restaurant Size",
       subtitle = "Modest relationship (non-linear) → Keep in model",
       x = "log(Seats)", y = "Lifespan (years)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", color = "#2c3e50"),
        plot.subtitle = element_text(color = "#f39c12"),
        panel.border = element_rect(color = "gray80", fill = NA, size = 0.8))

# Panel D: Foot Traffic (Standardized)
p_traffic <- ggplot(df_traffic_std, aes(x = foottraffic_std, y = lifespan_years)) +
  geom_point(alpha = 0.15, size = 1, color = "#95a5a6") +
  geom_smooth(method = "lm", color = "#e74c3c", fill = "#e74c3c", 
              alpha = 0.2, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", size = 0.5) +
  labs(title = "D. Foot Traffic (Pedestrian Count)",
       subtitle = "No meaningful relationship → EXCLUDED from model",
       x = "Pedestrians/day", y = "Lifespan (years)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", color = "#2c3e50"),
        plot.subtitle = element_text(color = "#e74c3c"),
        panel.border = element_rect(color = "gray80", fill = NA, size = 0.8))


# COMBINE PLOTS


p_final <- (p_same + p_other) / (p_size + p_traffic) +
  plot_annotation(
    title = "Feature comparison",
    subtitle = paste0("Analysis of ", nrow(df_standardized), 
                     " completed restaurants (unbiased sample) | All features standardized (μ=0, σ=1)"),
    theme = theme(
      plot.title = element_text(face = "bold", size = 18),
      plot.subtitle = element_text(size = 12, color = "gray30"),
      plot.caption = element_text(size = 8, color = "gray50", hjust = 0, lineheight = 1.2)
    )
  )


# SAVE PLOT


ggsave("feature_comparison_4panel_STANDARDIZED.png", p_final,
       width = 16, height = 12, dpi = 300, bg = "white")


# SUMMARY STATISTICS


summary_stats <- tibble(
  Feature = c("Same-Industry Competition", "Other-Industry Competition", 
              "Restaurant Size (log)", "Foot Traffic"),
  `Original Mean` = c(
    round(mean(df_standardized$same_industry_200m), 1),
    round(mean(df_standardized$other_industry_200m), 1),
    round(mean(df_standardized$log_seats), 2),
    round(mean(df_standardized$avg_daily_foottraffic, na.rm=TRUE), 0)
  ),
  `Original SD` = c(
    round(sd(df_standardized$same_industry_200m), 1),
    round(sd(df_standardized$other_industry_200m), 1),
    round(sd(df_standardized$log_seats), 2),
    round(sd(df_standardized$avg_daily_foottraffic, na.rm=TRUE), 0)
  ),
  `Std Mean` = c(
    round(mean(df_standardized$same_industry_std), 3),
    round(mean(df_standardized$other_industry_std), 3),
    round(mean(df_standardized$log_seats_std), 3),
    round(mean(df_standardized$foottraffic_std, na.rm=TRUE), 3)
  ),
  `Std SD` = c(
    round(sd(df_standardized$same_industry_std), 3),
    round(sd(df_standardized$other_industry_std), 3),
    round(sd(df_standardized$log_seats_std), 3),
    round(sd(df_standardized$foottraffic_std, na.rm=TRUE), 3)
  ),
  Correlation = c(cor_same, cor_other, cor_size, cor_traffic)
) %>%
  mutate(Correlation = round(Correlation, 3))

print(summary_stats)


# Display plot
print(p_final)

p_final
```





```{r}

library(tidyverse)
library(patchwork)

# Load & prep 
df_full <- readr::read_csv("df_ULTRA_enhanced_with_traffic.csv", show_col_types = FALSE)

df_std <- df_full %>%
  filter(bias_filter == FALSE, start_year_filter == TRUE)%>%              
  group_by(business_id) %>% slice_head(n = 1) %>% ungroup() %>%
  mutate(
    # total competition = same + other
    comp_total = coalesce(same_industry_200m, 0) + coalesce(other_industry_200m, 0),
    # median splits
    comp_level = if_else(
      comp_total > median(comp_total, na.rm = TRUE), "High Competition", "Low Competition",
      missing = NA_character_
    ),
    size_level = if_else(
      num_seats > median(num_seats, na.rm = TRUE), "Large (>median)", "Small (≤median)",
      missing = NA_character_
    )
  )

# enforce 2-level order and drop NA competition rows for any plot
df_std <- df_std %>%
  mutate(
    comp_level = factor(comp_level, levels = c("High Competition", "Low Competition")),
    size_level = factor(size_level, levels = c("Large (>median)", "Small (≤median)"))
  )

df_noNAcomp <- df_std %>% filter(!is.na(comp_level))

#  PANEL A: Competition × Size (unchanged design)
interactionA <- df_noNAcomp %>%
  filter(!is.na(size_level)) %>%
  group_by(comp_level, size_level) %>%
  summarise(
    n = n(),
    avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    .groups = "drop"
  )

p1_interact <- ggplot(interactionA, aes(x = comp_level, y = size_level, fill = avg_lifespan)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.1f yrs", avg_lifespan, n)),
            fontface = "bold", size = 4) +
  scale_fill_gradient2(low = "#e74c3c", mid = "#f39c12", high = "#27ae60",
                       midpoint = mean(df_noNAcomp$lifespan_years, na.rm = TRUE)) +
  labs(title = "A. Competition × Size Interaction",
       x = "Competition Level", y = "Restaurant Size", fill = "Avg Lifespan") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

#  Choose the two areas for Panels B/C 
area_pair <- c("Carlton", "Docklands")

df_area_pair <- df_noNAcomp %>%
  filter(clue_area %in% area_pair) %>%
  mutate(area_2 = factor(clue_area, levels = area_pair))   # lock order: Carlton, Docklands

#  PANEL B: Competition × Area (2×2, NO NA column) 
interactionB <- df_area_pair %>%
  group_by(comp_level, area_2) %>%
  summarise(
    n = n(),
    avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    .groups = "drop"
  )

p2_interact <- ggplot(interactionB, aes(x = comp_level, y = area_2, fill = avg_lifespan)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.1f yrs", avg_lifespan, n)),
            fontface = "bold", size = 4) +
  scale_fill_gradient2(low = "#e74c3c", mid = "#f39c12", high = "#27ae60",
                       midpoint = mean(df_area_pair$lifespan_years, na.rm = TRUE)) +
  labs(title = "B. Competition × Area (Carlton vs Docklands)",
       x = "Competition Level", y = "CLUE Area", fill = "Avg Lifespan") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

# PANEL C: Size × Area (2×2) 
interactionC <- df_area_pair %>%
  filter(!is.na(size_level)) %>%
  group_by(size_level, area_2) %>%
  summarise(
    n = n(),
    avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    .groups = "drop"
  )

p3_interact <- ggplot(interactionC, aes(x = size_level, y = area_2, fill = avg_lifespan)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.1f yrs", avg_lifespan, n)),
            fontface = "bold", size = 4) +
  scale_fill_gradient2(low = "#e74c3c", mid = "#f39c12", high = "#27ae60",
                       midpoint = mean(df_area_pair$lifespan_years, na.rm = TRUE)) +
  labs(title = "C. Size × Area (Carlton vs Docklands)",
       x = "Restaurant Size", y = "CLUE Area", fill = "Avg Lifespan") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

#  PANEL D: Does Size Buffer Competition? 
interactionD <- df_noNAcomp %>%
  filter(!is.na(size_level)) %>%
  group_by(comp_level, size_level) %>%
  summarise(
    avg_lifespan = mean(lifespan_years, na.rm = TRUE),
    .groups = "drop"
  )

p4_interact <- ggplot(interactionD, aes(x = comp_level, y = avg_lifespan,
                                        color = size_level, group = size_level)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  geom_text(aes(label = round(avg_lifespan, 1)),
            vjust = -1, fontface = "bold", size = 3.5, show.legend = FALSE) +
  scale_color_manual(values = c("Large (>median)" = "#27ae60", "Small (≤median)" = "#e74c3c")) +
  labs(title = "D. Does Size Buffer Competition?",
       x = "Competition Level", y = "Average Lifespan (years)", color = "Size") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

# save
p_interact_final <- (p1_interact | p2_interact) / (p3_interact | p4_interact) +
  plot_annotation(
    title = "Interaction Effects: Do Factors Work Together?",
    subtitle = "Panels B/C use Areas (Carlton vs Docklands). Competition = total competitors within 200m. Rows with missing competition are dropped.",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
                  plot.subtitle = element_text(size = 11, hjust = 0.5))
  )

ggsave("interaction_4panel_AREA_competition_noNA.png",
       p_interact_final, width = 14, height = 10, dpi = 300, bg = "white")

p_interact_final


```

R² ≈ 0.12 confirms that your models are exploratory, focusing on direction and relative strength of effects, not predictive accuracy.








```{r}



p1_lifecycle

p2_lifecycle

p3_lifecycle
```










































































































































































































































































